---
title: "Imputation of PEG genotype data to HRC reference panel"
output: html_notebook
---
```{r, setup, include=FALSE}
# We assume we're in the notebooks directory
basedir<-paste(getwd(),'/../rawdata/Genetics',sep='')
knitr::opts_knit$set(root.dir = basedir)
```
Install the latest version of:

Software for handling VCF files and indexing: [BCFtools](https://www.htslib.org/download/)

Software suite for GWAS: [PLINK2](https://www.cog-genomics.org/plink/2.0/)

Compression for bzip2 files: [7z](https://www.7-zip.org/download.html)

Prepare libraries in R that were referenced in PLINK2 tutorial:
```{r eval=FALSE}
packages <- c(
  "HardyWeinberg",
  "glue",
  "ggplot2",
  "plotly",
  "dplyr",
  "patchwork",
  "pROC",
  "ggforce",
  "data.table",
  "ggpubr",
  "ggrepel"
)

install_if_missing <- function(package) {
  if (!require(package, character.only = TRUE)) {
    cat(sprintf("Installing package: %s\n", package))  # Message indicating installation
    install.packages(package, dependencies = TRUE)
  } else {
    cat(sprintf("Package %s is already installed.\n", package))  # Message indicating already installed
  }
}

# Install the required packages
invisible(sapply(packages, install_if_missing))

# Confirm that installation is complete
cat("All packages checked and installed if missing.\n")
    
```

#source("../rawdata/R/loadplink2HW.R")
#source("../rawdata/R/popstrat.R")
#source("../rawdata/R/gwas1.R")

Download from the [Box folder](https://app.box.com/folder/344305501278) and save into **../rawdata/Genetics**:

- PEG_PD.phased.vcf.gz
- gwas_covariates.tar.gz

Convert the original VCF file into a PLINK2 formatted set of files:

```{bash eval=FALSE}
plink2 --vcf PEG_PD.phased.vcf.gz --make-pgen -out results/qc/qc1/PEG_PD_RAW
```

Then run the R script to generate a PLINK2 friendly pheno.txt by extracting data from the raw covariate files:

```{r eval=FALSE}
df_peg1_cov<-read.csv('peg1_covariates.csv')
df_peg1_cov<-df_peg1_cov[,c('ExternalDNACode','Female','Age','Ethnicity','PDstudyDiseaseNumeric')]
colnames(df_peg1_cov)<-c('PEGID','FEMALE','AGE','ETHNICITY','DISEASE')
df_peg1_cov$COHORT<-'PEG1'

df_peg2_cov<-read.csv('peg2_covariates.csv')
df_peg2_cov<-df_peg2_cov[,c('PEGID','Female','Age','EthnicityOriginal')]
df_peg2_cov$DISEASE<-1
colnames(df_peg2_cov)<-c('PEGID','FEMALE','AGE','ETHNICITY','DISEASE')
df_peg2_cov$COHORT<-'PEG2'

df_peg_all_cov<-rbind(df_peg1_cov,df_peg2_cov)
colnames(df_peg_all_cov)[4]<-'ETHNICITY'
df_peg_all_cov$DISEASE<-df_peg_all_cov$DISEASE+1

df_linkage<-read.csv('gwas_linkage.csv')
colnames(df_linkage)[1]<-'PEGID'

df_plink2_samples<-read.table('results/qc/qc1/PEG_PD_RAW.psam',sep='\t')
colnames(df_plink2_samples)<-c('GWAS_ID','SEX')
df_plink2_samples2<-merge(df_plink2_samples,df_linkage,by=c('GWAS_ID'),sort=FALSE)
df_plink2_samples3<-merge(df_plink2_samples2,df_peg_all_cov,by=c('PEGID'),sort=FALSE,all.x=TRUE)
df_plink2_samples3$FEMALE<-df_plink2_samples3$FEMALE+1
df_plink2_samples3$FEMALE[is.na(df_plink2_samples3$FEMALE)]<--9
df_plink2_samples3$AGE[is.na(df_plink2_samples3$AGE)]<--9
df_plink2_samples3$DISEASE[is.na(df_plink2_samples3$DISEASE)]<--9
df_plink2_samples3$ETHNICITY[df_plink2_samples3$ETHNICITY=='.']<-'Unknown'
df_plink2_samples3$ETHNICITY[is.na(df_plink2_samples3$ETHNICITY)]<-'NONE'
names(df_plink2_samples3)[which(names(df_plink2_samples3)=='GWAS_ID')]<-'IID'
write.table(df_plink2_samples3[,c('IID','ETHNICITY','DISEASE')],file='pheno.txt',quote=F,row.names=F)
write.table(df_plink2_samples3[,c('IID','FEMALE','AGE','COHORT')],file='covar.txt',quote=F,row.names=F)
```

Filter the dataset to include only subjects with disease status:

```{bash eval=FALSE}
plink2 --pfile results/qc/qc1/PEG_PD_RAW --pheno pheno.txt --make-pgen --require-pheno DISEASE -out results/qc/qc1/PEG_PD_COV_FILT
```

Compute Hardy-Weinberg stats:

```{bash eval=FALSE}
plink2 --pfile results/qc/qc1/PEG_PD_COV_FILT --hardy --snps-only --autosome --keep-founders --thin-indiv-count 100 --thin-count 100000 --out results/qc/qc1/PEG_PD_HW_ALL_POP
```
To get the HW Ternary plot in R that shows deviations across the 3 genotype classes:

```{r eval=FALSE}

# Load necessary package
library("HardyWeinberg")

# Function to load PLINK2 data into R
loadplink2Rmat <- function(file, header = TRUE, header_char = "#", sep = "\t", nrows = Inf) {
  # Read the header if it exists
  if (header) {
    header_line <- gsub(header_char, "", readLines(file, n = 1))  # Read and clean the header
    col_names <- strsplit(header_line, sep)[[1]]  # Split header into column names
    data <- read.table(file, sep = sep, skip = 1, header = FALSE, nrows = nrows)  # Read data, skipping header
    colnames(data) <- col_names  # Assign column names to the data
  } else {
    data <- read.table(file, sep = sep, header = FALSE, nrows = nrows)  # Read the entire file without skipping
    colnames(data) <- paste0("V", 1:ncol(data))  # Generate default column names
  }
  
  return(data)  # Return the loaded data
}

fname = "results/qc/qc1/PEG_PD_HW_ALL_POP.hardy"
source('../rawdata/R/loadplink2HW.R')
data<-loadplink2Rmat(fname)

HWTernaryPlot(data[,5:7],region=0,markercol = rgb(0,0,1,0.03),vertexlab = c("AA","AB","BB"))
```

QQ plot of HWE p-values:

```{r eval=FALSE}
format_hwdata2hwpack <- function(data) {
  x <- data[, 5:7]  # Select relevant columns
  colnames(x) <- c("AA", "AB", "BB")  # Set column names
  
  # Convert each column to numeric
  numeric_matrix <- do.call(cbind, lapply(x, function(column) as.numeric(unlist(column))))
  
  return(numeric_matrix)  # Return the numeric matrix
}
data2 <- format_hwdata2hwpack(data)
HWQqplot(data2,nsim=10,logplot=TRUE, main="Observed QQ HWE-Null")
```

See the allele frequency spectrum:

```{bash eval=FALSE}
plink2 --vcf PEG_PD.phased.vcf.gz \
	--freq alt1bins=0.01,0.02,0.03,0.04,0.05,0.06,0.07,0.08,0.09,0.10,0.11,0.12,0.13,0.14,0.15,0.16,0.17,0.18,0.19,0.20,0.21,0.22,0.23,0.24,0.25,0.26,0.27,0.28,0.29,0.30,0.31,0.32,0.33,0.34,0.35,0.36,0.37,0.38,0.39,0.40,0.41,0.42,0.43,0.44,0.45,0.46,0.47,0.48,0.49,0.50,0.51,0.52,0.53,0.54,0.55,0.56,0.57,0.58,0.59,0.60,0.61,0.62,0.63,0.64,0.65,0.66,0.67,0.68,0.69,0.70,0.71,0.72,0.73,0.74,0.75,0.76,0.77,0.78,0.79,0.80,0.81,0.82,0.83,0.84,0.85,0.86,0.87,0.88,0.89,0.90,0.91,0.92,0.93,0.94,0.95,0.96,0.97,0.98,0.99,1.00 \
	--snps-only \
	--autosome \
	--keep-founders \
	--thin-count 100000 \
	--out results/qc/qc1/PEG_PD_freq_ALLPOP_keepall
```

```{r}
fname='../rawdata/Genetics/results/qc/qc1/PEG_PD_freq_ALLPOP_keepal.afreq.alt1.bins'
data = loadplink2Rmat(fname)
barplot(unlist(data['OBS_CT']),names.arg=unlist(data['BIN_START']),ylim=c(0,1000),xlab="ALT1 FREQ",ylab="AUTOSOMAL SNP COUNT")
```

	
Remove SNPs with MAF <.01 call rates <.1 and out of HWE at p<1e-7
```{bash eval=FALSE}
cd ../rawdata/Genetics
plink2 --pfile results/qc/qc1/PEG_PD_COV_FILT --geno 0.1  --maf 0.01 -hwe 0.0000001 --export vcf --out results/qc/qc2/PEG_PD.phased.filtered
```


Generate Chr bgzip files

```{bash eval=FALSE}https://www.7-zip.org/download.html
cd ../rawdata/Genetics/results/qc/qc2
for chr in {1..22};
do
  echo Chrom $chr
  bcftools filter PEG_PD.phased.filtered.vcf.gz -r $chr > chunks/chr_$chr.vcf
  bgzip chunks/chr_$chr.vcf
  tabix chunks/chr_$chr.vcf.gz
done


```

Impute on [imputation server](https://imputationserver.sph.umich.edu/#!run/imputationserver2) after uploading these files.

Save results in ../rawdata/Genetics/imputed_raw:

```{bash eval=FALSE}
# replace password below with yours
pwd='vl8+UuqCENf>6P'
cd ../rawdata/Genetics/imputed_raw
for i in {1..22}
do
  echo "Chromosome $i"
  7z x -p$pwd chr_$i.zip
done

# Merge chr vcfs into a single one.
ls chr*.dose.vcf.gz >filelist.txt
bcftools concat -f filelist.txt -O z -o chr_all.dose.vcf.gz
```

Convert back to plink2:
```{bash eval=FALSE}
cd ../rawdata/Genetics
plink2 --vcf imputed_raw/chr_all.dose.vcf.gz dosage=DS --extract-if-info "R2 >= 0.8" --exclude-if-info "MAF < 0.01" --make-pgen --pheno pheno.txt -out results/qc/qc2/PEG_PD_IMPUTED
```

LD pruning:

```{bash eval=FALSE}
cd ../rawdata/Genetics
plink2 --pfile results/qc/qc2/PEG_PD_IMPUTED --rm-dup force-first --make-pgen --out results/qc/qc2/PEG_PD_dedup
plink2 --pfile results/qc/qc2/PEG_PD_dedup --indep-pairwise 100 0.8 --out results/qc/qc3/PEG_PD_prune
```


Generate new dataset of independent variants:

```{bash eval=FALSE}
cd ../rawdata/Genetics
plink2 --pfile results/qc/qc2/PEG_PD_dedup --make-pgen --extract results/qc/qc3/PEG_PD_prune.prune.in --out results/qc/qc3/PEG_PD_LE
```

Compute Principal Components:

```{bash eval=FALSE}
cd ../rawdata/Genetics
plink2 --pfile results/qc/qc3/PEG_PD_LE --freq counts --pca biallelic-var-wts --out results/qc/qc3/PEG_PD_pca
sed 's/^#//' results/qc/qc3/PEG_PD_LE.psam  > psam.tmp
sed 's/^#//' results/qc/qc3/PEG_PD_pca.eigenvec > evec.tmp
```
	
Generate PCA plots in R:

```{r eval=FALSE}
setwd('../rawdata/Genetics')
source('../R/popstrat.R')
popstrat_plot2d <- function(A, PCID_, Aprefix = NULL, A2 = NULL, PCID_A2 = NULL, A2prefix = NULL, column=NULL) {
  
  A$selected<-A[,which(colnames(A)==column)]
  A2$selected<-A2[,which(colnames(A2)==column)]
  
  # Negate values based on PCID_ for A
  for (i in 1:2) {
    if (grepl("^-", PCID_[i])) {
      negate_col <- substring(PCID_[i], 2)
      A[[PCID_[i]]] <- -A[[negate_col]]
    } 
  }
  
  A_names <- paste(Aprefix, A$selected)
  
  # Negate values for A2 if provided
  if (!is.null(A2) && !is.null(PCID_A2)) {
    for (i in 1:2) {
      if (grepl("^-", PCID_A2[i])) {
        negate_col <- substring(PCID_A2[i], 2)
        A2[[PCID_A2[i]]] <- -A2[[negate_col]]
      } 
    }
  }
  
  A2_names <- paste(A2prefix, A$selected)
  
  # Create the 2D scatter plot using plotly
  fig <- plot_ly() %>%
    add_trace(data = A, x = ~get(PCID_[1]), y = ~get(PCID_[2]), color = ~selected,
              colors = c('gold', 'red', 'green', 'blue', 'magenta'),
              type = "scatter", mode = "markers", 
              marker = list(size = 8, opacity = 1), 
              legendgroup = "A", name = A_names)
  
  # Add A2 data to the plot if provided
  if (!is.null(A2) && !is.null(PCID_A2)) {
    fig <- fig %>%
      add_trace(data = A2, x = ~get(PCID_A2[1]), y = ~get(PCID_A2[2]),
                type = "scatter", mode = "markers", color = ~selected, 
                colors = c('gold', 'red', 'green', 'blue', 'magenta'),
                marker = list(symbol = "square", size = 8, opacity = 1,
                              line = list(color = "black", width = 1)),
                legendgroup = "A2", name = A2_names)
  }
  
  fig <- fig %>%
    layout(
      xaxis = list(title = paste(Aprefix, PCID_[1], '|', A2prefix, PCID_A2[1])),
      yaxis = list(title = paste(Aprefix, PCID_[2], '|', A2prefix, PCID_A2[2])),
      showlegend = TRUE
    )
  
  # Print the plot
  show(fig)
}

df_eth<-read.table('../rawdata/Genetics/psam.tmp',sep='\t',header=TRUE)
df_pc<-read.table('../rawdata/Genetics/evec.tmp',sep='\t',header=TRUE)
df_merge<-merge(df_eth,df_pc,by=c('IID'))




popstrat_plot2d(df_merge,c('PC1','PC2'),Aprefix='ETHNICITY')

```

We see some outliers here, so let's filter out!
```{r eval=FALSE}
outliers<-df_pc[df_pc$PC1<  (-0.2),]
write.table(outliers$IID,file='../rawdata/Genetics/results/qc/qc3/pc1_outliers.txt',quote=F,row.names=F,col.names=F)	
```


Run second pass of Principal Components:

```{bash eval=FALSE}
cd ../rawdata/Genetics
plink2 --pfile results/qc/qc3/PEG_PD_LE --remove results/qc/qc3/pc1_outliers.txt --freq counts --pca biallelic-var-wts --out results/qc/qc3/PEG_PD_pca2
sed 's/^#//' results/qc/qc3/PEG_PD_pca2.eigenvec > evec.tmp
```

Generate new version with samples filtered out.

```{bash eval=FALSE}
cd ../rawdata/Genetics
plink2 --pfile results/qc/qc2/PEG_PD_IMPUTED --remove results/qc/qc3/pc1_outliers.txt --make-pgen --out results/qc/qc3/PEG_PD_IMPUTED2
```

Chunk into chromosomes:
```{bash eval=FALSE}
cd ../rawdata/Genetics
for i in {1..22}; do
 	  plink2 --pfile results/qc/qc3/PEG_PD_IMPUTED2 --chr $i --export Av --out chunks/PEG_PD_CHR_$i
done

```




Prepare for GWAS:

In R:

	df_covar<-read.table('covar.txt',header=T)
	df_eth<-read.table('evec.tmp',header=T,sep='\t')
	df_merge<-merge(df_covar,df_eth,by=c('IID'))
	write.table(df_merge,file='covar_evec.txt',row.names=F,quote=F)

Run GWAS:

	plink2 --pfile results/qc/qc3/PEG_PD_IMPUTED2 --glm omit-ref hide-covar no-firth --pheno-name DISEASE --covar covar_evec.txt --covar-name AGE,FEMALE,PC1,PC2 --vif 1000 --covar-variance-standardize -pfilter 1 --out results/gwas/PEG_PD
	sed 's/^#//' results/gwas/PEG_PD.DISEASE.glm.logistic > ../GWAS_meta/gwas_peg.txt

Prep for meta-analysis in R:

	setwd('[repo_root]/rawdata/Genetics/')
	df_peg<-read.table('GWAS_meta/gwas.txt',sep='\t',header=T)	
	df_nalls<-read.table('../Nalls_GWAS/PDNallsP005.csv',header=T,sep=',')
	df_nalls$ID<-paste(df_nalls$chr,df_nalls$position,sep=':')
	df_nalls$N<-df_nalls$ncase+df_nalls$ncontrol
	write.csv(df_nalls,file='../GWAS_meta/gwas_nalls.txt',row.names=F,quote=F)
	df_merged<-merge(df_peg,df_nalls,by.x=c('CHROM','POS'),by.y=c('chr','position'))

Run meta in [repo_root]/rawdata/GWAS_meta:
	../metal < metal.script

Sort:
	df_meta<-read.table('GWAS_OUT1.metal',sep='\t',header=T)
	write.table(ordered,file='sorted_meta.txt',quote=F,row.names=F)


Make a dataframe file from plink2:

	plink2 --vcf PEG_PD.phased.filtered.imputed.vcf.gz --export Av --out PEG_PD.dataframe

Post-process this dataframe to split into two files:

	./process_plink_dataframe.py < PEG_PD.dataframe.traw

rewrite the header based on mapping files to prepare for Matrix EQTL:

	rewrite_header.py mapping_peg.csv ',' PEG_PD.geno.txt ' ' > ../MatrixEQTL/mapped_peg_by_snp.txt

In MatrixEQTL make symlinks
In <repo_root>/rawdata/MatrixEQTL/peg1 for example:

	ln -s ../mapped_peg_by_snp.txt

## Intersect columns:

	intersect_by_header.py ' ' peg1_ancestry_by_covariate.txt mapped_peg_by_snp.txt mapped_peg1_by_cpg.txt

Filter on DMPs

	filter_on_whitelist.py ../dmp_cpgs.txt  ' ' < intersected_mapped_peg1_by_cpg.txt > intersected_mapped_peg1_by_cpg_dmp.txt

