---
title: "mQTL Meta-Analysis: PEG + PPMI"
author: "Your Name"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    code_folding: show
  pdf_document:
    toc: true
    toc_depth: '3'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = NA,
  warning = FALSE,
  cache   = TRUE,
  fig.width  = 8,
  fig.height = 6
)
options(knitr.chunk.echo=TRUE)
```

```{r libraries}
suppressPackageStartupMessages({
  library(tidyverse)
  library(data.table)
  library(meta)        # r-meta: metagen() for inverse-variance meta-analysis
  library(ggplot2)
  library(patchwork)
  library(scales)
})
```

---

# 1. Configuration

```{r config}
# ── Paths to cis mQTL result files ───────────────────────────────────────────
RESULTS <- list(
  peg1_cases    = "../../results/PEG_MQTL/mqtl/peg1_cases/cis_mqtls.txt",
  peg1_controls = "../../results/PEG_MQTL/mqtl/peg1_controls/cis_mqtls.txt",
  peg2_cases    = "../../results/PEG_MQTL/mqtl/peg2_cases/cis_mqtls.txt",
  ppmi_cases    = "../../results/PPMI_MQTL/mqtl/ppmi_cases/cis_mqtls.txt",
  ppmi_controls = "../../results/PPMI_MQTL/mqtl/ppmi_controls/cis_mqtls.txt"
)

# Row cap per cohort file. The per-cohort cis thresholds are deliberately
# liberal (1e-2 for PEG_MQTL, 1e-3 for PPMI_MQTL) so that enough pairs survive
# for the meta-analysis to combine evidence across cohorts — which makes the
# raw files far too large to read whole on this machine.
#
# This cap is NOT the same one the overlap report applies: that one bounds the
# PEG_METAB trans metQTL files, which this report never reads. Nothing caps
# these files at write time, so this is the only guard on the cis path.
#
# Note this is a real analytic trade-off, not just a memory guard: truncation
# happens before cohorts are pooled, so a pair that would have reached
# significance only by combining weak-but-concordant evidence across cohorts
# can be cut. The load step warns when a file actually hits the cap.
MAX_CIS_ROWS <- 10000000

# PEG_MQTL and PPMI_MQTL now write gzipped output (matrixeqtl.gzip_output in
# their config.yaml). Prefer the .gz when it exists so this report works both
# before and after those pipelines are re-run, without editing paths twice.
resolve_path <- function(p) if (file.exists(paste0(p, ".gz"))) paste0(p, ".gz") else p
RESULTS <- lapply(RESULTS, resolve_path)

# Meta-analysis groupings
# Cases:    PEG1 × PPMI two-way — comparable sample sizes to controls
#           PEG2 excluded from primary; used as sensitivity analysis below
# Controls: PEG1 × PPMI two-way
# Rationale: 3-way PEG1 ∩ PEG2 ∩ PPMI yields only ~2 pairs;
#            PEG1 ∩ PPMI yields ~1032 pairs which is tractable.
META_GROUPS <- list(
  cases    = c("peg1_cases",    "ppmi_cases"),
  controls = c("peg1_controls", "ppmi_controls")
)

# Sensitivity: 3-way intersection including PEG2 cases
SENSITIVITY_GROUPS <- list(
  cases_peg2_sensitivity = c("peg1_cases", "peg2_cases", "ppmi_cases")
)

# Output directory
META_DIR <- "../../results/META_MQTL/meta"
dir.create(META_DIR, recursive=TRUE, showWarnings=FALSE)
dir.create("../../results/META_MQTL/qc", recursive=TRUE, showWarnings=FALSE)

# Labels and colors
DATASET_LABELS <- c(
  peg1_cases    = "PEG1 Cases",
  peg1_controls = "PEG1 Controls",
  peg2_cases    = "PEG2 Cases",
  ppmi_cases    = "PPMI Cases",
  ppmi_controls = "PPMI Controls"
)

DATASET_COLORS <- c(
  peg1_cases    = "#E41A1C",
  peg1_controls = "#377EB8",
  peg2_cases    = "#FF7F00",
  ppmi_cases    = "#4DAF4A",
  ppmi_controls = "#984EA3"
)
```

---

# 2. Load mQTL Results

```{r load-results, cache=TRUE, cache.lazy = FALSE}
load_mqtl <- function(path, dataset_name) {
  message(sprintf("Loading dataset %s",dataset_name))
  if (!file.exists(path)) {
    warning(sprintf("File not found: %s", path))
    return(NULL)
  }
  # MatrixEQTL writes p-value ascending when it computes FDR
  # (noFDRsaveMemory = FALSE), so capping keeps the most significant rows and
  # discards only the tail. See MAX_CIS_ROWS above for the trade-off.
  #
  # Read via `zcat | head` rather than fread(nrows=) on the gzip directly:
  # fread decompresses the *entire* file before honouring nrows (measured: read
  # time scales with total file size, 1.8s -> 5.4s for a 4x larger file at the
  # same cap), which on a multi-GB cis file means expanding every row we are
  # about to discard. Piping through head stops at the cap — constant time and
  # the lowest peak memory of the options tested.
  if (grepl("\\.gz$", path)) {
    dat <- fread(cmd = sprintf("zcat %s | head -n %.0f",
                               shQuote(path), MAX_CIS_ROWS + 1),
                 data.table = FALSE)
  } else {
    dat <- fread(path, data.table = FALSE, nrows = MAX_CIS_ROWS)
  }

  # A failed decompress or a bad path surfaces as an empty read rather than an
  # error, which would otherwise sail past the file.exists() check above and
  # silently drop a whole cohort from the meta-analysis.
  if (nrow(dat) == 0) {
    warning(sprintf("%s: read 0 rows from %s — decompression or path problem?",
                    dataset_name, path))
    return(NULL)
  }

  # Reported with cat(), not warning(): this document sets warning = FALSE in
  # its global chunk options, so a warning() here is swallowed outright and the
  # truncation would go unrecorded. Truncation changes which pairs can reach
  # significance, so it belongs in the rendered output, not the console.
  if (nrow(dat) == MAX_CIS_ROWS) {
    pcol <- intersect(c("p-value", "p.value", "pvalue"), colnames(dat))
    worst_p <- if (length(pcol) > 0) max(dat[[pcol[1]]], na.rm=TRUE) else NA_real_
    cat(sprintf(
      paste0("  *** %s: HIT THE %s-ROW CAP — least significant rows dropped ",
             "before meta-analysis (worst p retained: %.3g). Pairs that would ",
             "only reach significance by pooling across cohorts may be lost. ",
             "Raise MAX_CIS_ROWS if RAM allows. ***\n"),
      dataset_name, format(MAX_CIS_ROWS, big.mark=","), worst_p))
  }

  # Normalise column names — MatrixEQTL output varies between pipelines:
  # PEG uses:  SNP  gene  beta  t-stat  p-value  FDR
  # PPMI uses: snps gene  beta  t-stat  p-value  FDR
  col_map <- c(
    "SNP"     = "snp_id",
    "snps"    = "snp_id",
    "gene"    = "cpg_id",
    "p-value" = "pvalue",
    "p.value" = "pvalue",
    "t-stat"  = "tstat",
    "t.stat"  = "tstat",
    "FDR"     = "fdr_dataset"
  )
  matches <- col_map[names(col_map) %in% colnames(dat)]
  if (length(matches) > 0)
    setnames(dat, old=names(matches), new=unname(matches))

  required <- c("snp_id", "cpg_id", "beta", "pvalue", "tstat")
  missing  <- setdiff(required, colnames(dat))
  if (length(missing) > 0) {
    warning(sprintf(
      "%s: missing columns after rename: %s\nActual columns: %s",
      dataset_name, paste(missing, collapse=", "),
      paste(colnames(dat), collapse=", ")
    ))
    return(NULL)
  }

  # ── Normalise SNP IDs to chr:pos so pair_id matches across cohorts ─────────
  # PEG SNP IDs are already chr:pos (e.g. "1:1446221") from imputation.
  # PPMI SNP IDs are chr:pos:ref:alt (e.g. "1:1365759:G:A") from WGS.
  # Stripping any trailing :REF:ALT normalises both to a common chr:pos key
  # without needing a lookup join, which avoids brittleness if the position
  # files and mQTL output ever drift out of sync.
  dat$snp_chrpos <- sub("^([0-9XYM]+:[0-9]+):[ACGT]+:[ACGT]+$", "\\1", dat$snp_id)

  n_stripped <- sum(dat$snp_chrpos != dat$snp_id)
  cat(sprintf("  %-20s: %s SNP IDs had allele suffix stripped (%s unchanged)\n",
      dataset_name,
      format(n_stripped, big.mark=","),
      format(nrow(dat) - n_stripped, big.mark=",")))

  dat <- dat %>%
    mutate(
      dataset = dataset_name,
      se      = abs(beta / tstat),
      pair_id = paste(snp_chrpos, cpg_id, sep="|")
    ) %>%
    filter(!is.na(se), se > 0, !is.na(beta), !is.na(pvalue))

  cat(sprintf("  %-20s: %s associations\n",
      dataset_name, format(nrow(dat), big.mark=",")))
  dat
}


cat("Loading cis mQTL results:\n")
message("Loading cis mQTL results:\n")
mqtl_list <- imap(RESULTS, function(path, nm) {
  load_mqtl(path, nm)
})
mqtl_list <- Filter(Negate(is.null), mqtl_list)

cat(sprintf("\nDatasets loaded: %d / %d\n",
    length(mqtl_list), length(RESULTS)))
```

---

# 3. P-value QC

## 3.1 Result Summary

> **Note on Lambda GC:** Lambda GC is not reported here because MatrixEQTL
> outputs only associations passing the significance threshold
> (`pval_cis = 1e-3` for PEG, `pval_cis = 1e-5` for PPMI). Computing lambda
> on a filtered tail of the p-value distribution produces artificially
> inflated values that are meaningless. Lambda GC is only valid when computed
> on the complete null distribution. QQ plots in Section 3.2 provide valid
> visual QC for the significant associations we do have.

```{r result-summary}
summary_df <- imap_dfr(mqtl_list, function(dat, nm) {
  data.frame(
    dataset     = nm,
    label       = DATASET_LABELS[nm],
    n_assoc     = nrow(dat),
    n_fdr05     = if ("fdr_dataset" %in% colnames(dat))
                    sum(dat$fdr_dataset < 0.05, na.rm=TRUE) else NA,
    n_p1e5      = sum(dat$pvalue < 1e-5,  na.rm=TRUE),
    n_p1e8      = sum(dat$pvalue < 1e-8,  na.rm=TRUE),
    median_beta = round(median(abs(dat$beta), na.rm=TRUE), 4),
    max_beta    = round(max(abs(dat$beta),    na.rm=TRUE), 4),
    n_extreme   = sum(abs(dat$beta) > 1,   na.rm=TRUE),
    stringsAsFactors = FALSE
  )
})

summary_df %>%
  mutate(
    n_assoc = format(n_assoc, big.mark=","),
    n_fdr05 = format(n_fdr05, big.mark=","),
    n_p1e5  = format(n_p1e5,  big.mark=","),
    n_p1e8  = format(n_p1e8,  big.mark=",")
  ) %>%
  select(label, n_assoc, n_fdr05, n_p1e5, n_p1e8,
         median_beta, max_beta, n_extreme) %>%
  knitr::kable(
    col.names = c("Dataset", "N assoc", "FDR<0.05", "p<1e-5", "p<1e-8",
                  "Median |β|", "Max |β|", "N |β|>1 (outliers)"),
    caption   = paste("Per-dataset cis-mQTL summary.",
                      "FDR computed by MatrixEQTL over all tested cis pairs",
                      "(noFDRsaveMemory=FALSE) — valid for within-dataset use.",
                      "Cross-cohort FDR (FDR_BH) is computed on meta-analysis",
                      "p-values in Section 5.")
  )

# ── Flag extreme beta values ──────────────────────────────────────────────────
# Methylation beta values should be bounded 0-1. Effect sizes (beta) in mQTL
# analysis represent change in methylation per unit allele dosage and should
# therefore be well within [-1, 1] for a properly scaled analysis.
# Values >> 1 indicate a scaling issue in the methylation or covariate matrix.
extreme_datasets <- summary_df %>% filter(max_beta > 1)
if (nrow(extreme_datasets) > 0) {
  cat("\n⚠️  WARNING: Extreme beta values detected (|β| > 1):\n")
  for (i in seq_len(nrow(extreme_datasets))) {
    cat(sprintf("   %-20s: max|β| = %.2f, n outliers = %s\n",
                extreme_datasets$dataset[i],
                extreme_datasets$max_beta[i],
                format(extreme_datasets$n_extreme[i], big.mark=",")))
  }
  cat("\n   This most likely indicates a scaling issue in the PPMI methylation\n")
  cat("   or covariate matrix. Possible causes:\n")
  cat("   - Methylation matrix not in beta [0,1] scale (e.g. M-values used)\n")
  cat("   - A covariate column with large numeric range (e.g. age in years\n")
  cat("     not standardised) creating numerical instability in MatrixEQTL\n")
  cat("   - Batch variable included without dummy-coding\n")
  cat("   Recommend investigating PPMI results before proceeding with meta-analysis.\n")
}
```

```{r counts-plot, fig.height=4}
# Bar chart of association counts at different p-value thresholds
summary_df %>%
  select(label, dataset, n_p1e5, n_p1e8) %>%
  pivot_longer(c(n_p1e5, n_p1e8),
               names_to="threshold", values_to="n") %>%
  mutate(threshold = factor(threshold,
    levels = c("n_p1e5", "n_p1e8"),
    labels = c("p < 1e-5", "p < 1e-8"))) %>%
  ggplot(aes(x=label, y=n, fill=dataset)) +
  geom_col(position="dodge", width=0.7) +
  scale_fill_manual(values=DATASET_COLORS, guide="none") +
  scale_y_continuous(labels=comma) +
  facet_wrap(~threshold, scales="free_y") +
  labs(title="Significant Associations per Dataset",
       x=NULL, y="Count") +
  theme_minimal(base_size=11) +
  theme(axis.text.x=element_text(angle=30, hjust=1))
```

## 3.2 QQ Plots

```{r qq-function}
# n_total: true number of tests the retained p-values were drawn from
# (e.g. all cis SNP-CpG pairs MatrixEQTL actually evaluated), when the
# input pvalues are only the smallest n of a much larger, left-truncated
# set. Order statistic i of n retained values is then the i-th smallest
# of n_total, not of n, so expected quantiles and the Beta CI ribbon
# must be computed against n_total. Defaults to n (i.e. assumes pvalues
# is the complete, untruncated set) when n_total is not supplied.
make_qq <- function(pvalues, title, color="steelblue", max_pts=1e6, n_total=NULL) {
  pvalues <- pvalues[!is.na(pvalues) & pvalues > 0 & pvalues <= 1]
  top     <- pvalues <= 1e-4
  if (sum(!top) > max_pts)
    pvalues <- c(pvalues[top],
                 sample(pvalues[!top], max_pts))
  n   <- length(pvalues)
  if (is.null(n_total)) n_total <- n
  rank  <- seq_len(n)
  a     <- 0.5
  obs   <- sort(-log10(pvalues))
  exp   <- sort(-log10((rank - a) / (n_total + 1 - 2 * a)))
  ci_hi <- -log10(qbeta(0.025, rank, n_total - rank + 1))
  ci_lo <- -log10(qbeta(0.975, rank, n_total - rank + 1))

  ggplot(data.frame(exp=exp, obs=obs,
                    ci_hi=sort(ci_hi), ci_lo=sort(ci_lo)),
         aes(x=exp, y=obs)) +
    geom_ribbon(aes(ymin=ci_lo, ymax=ci_hi),
                fill="lightblue", alpha=0.4) +
    geom_point(size=0.5, alpha=0.5, color=color) +
    geom_abline(slope=1, intercept=0,
                color="red", linewidth=0.7) +
    labs(title    = title,
         subtitle = sprintf("n = %s%s", format(n, big.mark=","),
                             if (n_total > n)
                               sprintf(" (of %s cis tests)", format(n_total, big.mark=","))
                             else ""),
         x="Expected  –log₁₀(p)",
         y="Observed  –log₁₀(p)") +
    theme_minimal(base_size=11) +
    theme(plot.subtitle=element_text(size=9, color="grey40"))
}
```

```{r qq-plots, fig.width=14, fig.height=9, cache=TRUE}
n_cis_tests_df <- read.csv("../../results/META_MQTL/qc/n_cis_tests.csv")
n_cis_tests <- setNames(n_cis_tests_df$n_total, n_cis_tests_df$dataset)

qq_plots <- imap(mqtl_list, function(dat, nm)
  make_qq(dat$pvalue, DATASET_LABELS[nm], DATASET_COLORS[nm],
          n_total = n_cis_tests[[nm]]))

wrap_plots(qq_plots, ncol=3) +
  plot_annotation(
    title = "QQ Plots — Per-Dataset Cis mQTL Analyses",
    theme = theme(plot.title=element_text(size=15, face="bold"))
  )
```

## 3.3 Effect Size Distributions

```{r beta-dist, fig.height=4}
imap_dfr(mqtl_list, function(dat, nm)
  dat %>%
    slice_sample(n=min(100000, nrow(dat))) %>%
    mutate(label=DATASET_LABELS[nm])
) %>%
  ggplot(aes(x=beta, color=dataset, fill=dataset)) +
  geom_density(alpha=0.15, linewidth=0.7) +
  scale_color_manual(values=DATASET_COLORS, labels=DATASET_LABELS) +
  scale_fill_manual( values=DATASET_COLORS, labels=DATASET_LABELS) +
  geom_vline(xintercept=0, linetype="dashed", color="grey40") +
  labs(title="Effect Size (β) Distributions — Cis mQTLs",
       subtitle="Sampled to 100K associations per dataset",
       x="Beta", y="Density", color=NULL, fill=NULL) +
  theme_minimal(base_size=13) +
  theme(legend.position="bottom")
```

---

# 4. Meta-Analysis with r-meta

## 4.1 Strategy

We use **fixed-effects inverse-variance weighted (IVW) meta-analysis**
via `metagen()` from the `meta` package. Fixed effects is appropriate
here because we are combining results from cohorts drawn from the same
biological population — we want to estimate a common true effect, not
the distribution of effects across populations.

For each SNP-CpG pair tested in multiple datasets:

- **Effect size:** β from MatrixEQTL
- **Standard error:** SE = |β / t-stat|
- **Method:** IVW fixed effects (`method.tau="DL"` for heterogeneity)
- **Heterogeneity:** Cochran's Q and I² reported for significant hits

## 4.2 Run Meta-Analysis Per Group

```{r meta-analysis, cache=TRUE}
run_meta_group <- function(group_name, dataset_names) {
  cat(sprintf("\n=== %s meta-analysis ===\n", group_name))
  message(sprintf("\n=== %s meta-analysis ===\n", group_name))
  group_data <- mqtl_list[dataset_names]
  group_data <- Filter(Negate(is.null), group_data)

  if (length(group_data) == 0) {
    cat("No data available\n")
    return(NULL)
  }

  cat(sprintf("Datasets: %s\n", paste(names(group_data), collapse=", ")))
  message(sprintf("Datasets: %s\n", paste(names(group_data), collapse=", ")))
  # ── Strict intersection ───────────────────────────────────────────────────
  # Only meta-analyse pairs significant in ALL constituent datasets.
  # This is computationally tractable and focuses on robustly replicated signals.
  # Individual cohort thresholds are set permissively (pval_cis = 1e-3 for PEG,
  # 1e-5 for PPMI) so the intersection captures pairs that clear the bar in both.
  common_pairs <- Reduce(intersect, lapply(group_data, function(d) d$pair_id))
  cat(sprintf("SNP-CpG pairs in all datasets: %s\n",
      format(length(common_pairs), big.mark=",")))

  message(sprintf("SNP-CpG pairs in all datasets: %s\n",
      format(length(common_pairs), big.mark=",")))
  if (length(common_pairs) == 0) {
    cat("No common pairs — check SNP and probe intersection\n")
    return(NULL)
  }

  # Subset each dataset to common pairs and align
  aligned <- lapply(names(group_data), function(nm) {
    group_data[[nm]] %>%
      filter(pair_id %in% common_pairs) %>%
      select(pair_id, snp_id, cpg_id, beta, se, pvalue) %>%
      rename_with(~paste0(., "_", nm), c(beta, se, pvalue, snp_id, cpg_id))
  })

  # Inner join — all pairs present in all datasets by construction
  merged <- Reduce(function(a, b) inner_join(a, b, by="pair_id"), aligned)

  # Recover snp_id and cpg_id from first dataset
  first_nm <- names(group_data)[1]
  merged <- merged %>%
    rename(snp_id = paste0("snp_id_", first_nm),
           cpg_id = paste0("cpg_id_", first_nm)) %>%
    select(-matches("^snp_id_|^cpg_id_"))

  cat(sprintf("Pairs after alignment join: %s\n",
      format(nrow(merged), big.mark=",")))
  message(sprintf("Pairs after alignment join: %s\n",
      format(nrow(merged), big.mark=",")))

  beta_cols <- paste0("beta_", names(group_data))
  se_cols   <- paste0("se_",   names(group_data))

  cat("Running metagen()...\n")

  message("Running metagen()...\n")

  meta_results <- lapply(seq_len(nrow(merged)), function(i) {
    betas <- as.numeric(merged[i, beta_cols])
    ses   <- as.numeric(merged[i, se_cols])

    if (any(is.na(betas)) || any(is.na(ses)) || any(ses <= 0))
      return(NULL)

    mg <- tryCatch(
      metagen(TE=betas, seTE=ses,
              studlab    = names(group_data),
              method.tau = "DL",
              common     = TRUE,
              random     = FALSE),
      error = function(e) NULL
    )
    if (is.null(mg)) return(NULL)

    data.frame(
      pair_id    = merged$pair_id[i],
      snp_id     = merged$snp_id[i],
      cpg_id     = merged$cpg_id[i],
      meta_beta  = mg$TE.common,
      meta_se    = mg$seTE.common,
      meta_pval  = mg$pval.common,
      meta_z     = mg$zval.common,
      Q          = mg$Q,
      Q_pval     = mg$pval.Q,
      I2         = mg$I2,
      n_datasets = length(betas),
      stringsAsFactors = FALSE
    )
  })

  meta_df <- bind_rows(Filter(Negate(is.null), meta_results))

  if (nrow(meta_df) == 0) {
    cat("No results produced\n")
    return(NULL)
  }

  # FDR correction
  meta_df <- meta_df %>%
    arrange(meta_pval) %>%
    mutate(FDR_BH = p.adjust(meta_pval, method="BH"))

  cat(sprintf("Meta-analysis complete: %s pairs\n",
      format(nrow(meta_df), big.mark=",")))
  cat(sprintf("  FDR < 0.05: %s\n",
      format(sum(meta_df$FDR_BH < 0.05, na.rm=TRUE), big.mark=",")))
  cat(sprintf("  FDR < 0.10: %s\n",
      format(sum(meta_df$FDR_BH < 0.10, na.rm=TRUE), big.mark=",")))
  cat(sprintf("  p < 1e-8:   %s\n",
      format(sum(meta_df$meta_pval < 1e-8, na.rm=TRUE), big.mark=",")))

  # Save full results
  out_file <- file.path(META_DIR,
                         paste0(group_name, "_meta_results.csv.gz"))
  fwrite(meta_df, out_file)
  cat(sprintf("  Saved: %s\n", basename(out_file)))

  invisible(meta_df)
}

# NOTE: metagen() is called once per SNP-CpG pair.
# For large datasets this can take 30-60 minutes.
# Results are cached — delete cache/ to rerun.
meta_results <- list()
for (group in names(META_GROUPS)) {
  meta_results[[group]] <- run_meta_group(group, META_GROUPS[[group]])
}

# ── Sensitivity analysis: 3-way intersection including PEG2 cases ─────────────
# PEG2 is cases-only (Wave 2). The 3-way PEG1 ∩ PEG2 ∩ PPMI intersection
# is very strict but identifies the most robustly replicated signals.
message("Running sensitivity analysis (PEG1 + PEG2 + PPMI cases 3-way)...")
sensitivity_results <- list()
for (group in names(SENSITIVITY_GROUPS)) {
  sensitivity_results[[group]] <- run_meta_group(
    group, SENSITIVITY_GROUPS[[group]])
}

# ── Flag primary cases hits also significant in PEG2 ─────────────────────────
if (!is.null(meta_results$cases) && nrow(meta_results$cases) > 0) {
  peg2_pairs <- mqtl_list$peg2_cases$pair_id
  meta_results$cases$in_peg2 <- meta_results$cases$pair_id %in% peg2_pairs
  message(sprintf("Primary cases hits also significant in PEG2: %d / %d",
      sum(meta_results$cases$in_peg2, na.rm=TRUE),
      nrow(meta_results$cases)))
}
```

---

# 5. Meta-Analysis Results

## 5.1 Summary

```{r meta-summary}
summary_df <- imap_dfr(meta_results, function(dat, group) {
  if (is.null(dat) || nrow(dat) == 0) return(NULL)
  data.frame(
    group      = str_to_title(group),
    n_pairs    = nrow(dat),
    pct_fdr05  = round(100 * mean(dat$FDR_BH < 0.05, na.rm=TRUE), 1),
    n_fdr05    = sum(dat$FDR_BH    < 0.05, na.rm=TRUE),
    n_fdr10    = sum(dat$FDR_BH    < 0.10, na.rm=TRUE),
    n_p1e5     = sum(dat$meta_pval < 1e-5, na.rm=TRUE),
    n_p1e8     = sum(dat$meta_pval < 1e-8, na.rm=TRUE),
    n_in_peg2  = if ("in_peg2" %in% colnames(dat))
                   sum(dat$in_peg2, na.rm=TRUE) else NA,
    pct_het    = round(100 * mean(dat$Q_pval < 0.05, na.rm=TRUE), 1),
    stringsAsFactors = FALSE
  )
})

if (is.null(summary_df) || nrow(summary_df) == 0) {
  cat("No meta-analysis results available yet — run Section 4.2 first.
")
} else {
  summary_df %>%
    mutate(across(c(n_pairs, n_fdr05, n_fdr10, n_p1e5, n_p1e8, n_in_peg2),
                  ~format(.x, big.mark=","))) %>%
    knitr::kable(
      col.names = c("Group", "Pairs", "% FDR<0.05", "FDR<0.05",
                    "FDR<0.10", "p<1e-5", "p<1e-8",
                    "Also in PEG2", "% heterogeneous"),
      caption   = paste("Meta-analysis summary (Cases: PEG1 × PPMI two-way;",
                        "Controls: PEG1 × PPMI two-way).",
                        "'Also in PEG2' = primary hits also significant in PEG2",
                        "at pval_cis < 1e-3, providing three-cohort support.",
                        "FDR_BH computed on meta-analysis p-values.")
    )
    
}
```

## 5.2 QQ Plots — Meta-Analysis

```{r meta-qq, fig.width=11, fig.height=5}
meta_qq <- imap(meta_results, function(dat, group) {
  if (is.null(dat)) return(NULL)
  make_qq(dat$meta_pval,
          title = sprintf("%s Meta-Analysis", str_to_title(group)),
          color = ifelse(group=="cases", "#E41A1C", "#377EB8"))
})
meta_qq <- Filter(Negate(is.null), meta_qq)

if (length(meta_qq) > 0)
  wrap_plots(meta_qq, ncol=2) +
    plot_annotation(
      title = "QQ Plots — Meta-Analysis Results",
      theme = theme(plot.title=element_text(size=15, face="bold"))
    )
```

## 5.3 Top Meta-mQTLs

```{r top-meta-hits}
fmt_pval <- function(x) {
  case_when(
    is.na(x)       ~ "NA",
    x < 1e-300     ~ "< 1e-300",
    x < 0.001      ~ formatC(x, format="e", digits=2),
    TRUE           ~ as.character(round(x, 4))
  )
}

for (group in names(meta_results)) {
  dat <- meta_results[[group]]
  if (is.null(dat) || nrow(dat) == 0) {
    cat(sprintf("\n### %s — no results\n", str_to_title(group)))
    next
  }

  cat(sprintf("\n### %s — Top 20\n", str_to_title(group)))

  top20 <- dat %>%
    arrange(meta_pval) %>%
    slice_head(n=20)

  # Pull per-study beta and p-value for each top pair
  datasets_in_group <- META_GROUPS[[group]]
  per_study <- lapply(datasets_in_group, function(nm) {
    mqtl_list[[nm]] %>%
      filter(pair_id %in% top20$pair_id) %>%
      select(pair_id, beta, pvalue) %>%
      rename_with(~paste0(nm, "_", .), c(beta, pvalue))
  })
  per_study_df <- Reduce(function(a, b)
    left_join(a, b, by="pair_id"), per_study)

  # Join per-study stats to top20
  top20 <- top20 %>%
    left_join(per_study_df, by="pair_id")

  # Build dynamic column list
  study_beta_cols <- paste0(datasets_in_group, "_beta")
  study_pval_cols <- paste0(datasets_in_group, "_pvalue")

  top20 %>%
    select(snp_id, cpg_id, meta_beta, meta_se,
           meta_pval, FDR_BH, I2, Q_pval,
           all_of(c(study_beta_cols, study_pval_cols))) %>%
    mutate(
      meta_beta = round(meta_beta, 4),
      meta_se   = round(meta_se,   4),
      meta_pval = fmt_pval(meta_pval),
      FDR_BH    = fmt_pval(FDR_BH),
      I2        = round(I2, 1),
      Q_pval    = fmt_pval(Q_pval),
      across(all_of(study_beta_cols), ~round(.x, 4)),
      across(all_of(study_pval_cols), ~fmt_pval(.x))
    ) %>%
    knitr::kable(
      col.names = c(
        "SNP", "CpG", "Meta β", "Meta SE",
        "Meta p", "FDR", "I²", "Q p",
        paste0(datasets_in_group, " β"),
        paste0(datasets_in_group, " p")
      ),
      caption = sprintf("Top 20 %s meta-mQTLs with per-study statistics",
                        group)
    ) %>%
    print()
}
```

## 5.4 Cases vs Controls Comparison

For mQTL pairs significant in both meta-analyses, compare effect sizes:

```{r cases-vs-controls, fig.height=5}
if (!is.null(meta_results$cases) && !is.null(meta_results$controls)) {

  shared <- inner_join(
    meta_results$cases    %>% select(pair_id, snp_id, cpg_id,
                                      beta_cases=meta_beta,
                                      pval_cases=meta_pval,
                                      fdr_cases=FDR_BH),
    meta_results$controls %>% select(pair_id,
                                      beta_ctrls=meta_beta,
                                      pval_ctrls=meta_pval,
                                      fdr_ctrls=FDR_BH),
    by="pair_id"
  )

  cat(sprintf("Pairs tested in both groups: %s\n",
      format(nrow(shared), big.mark=",")))

  sig_either <- shared %>%
    filter(fdr_cases < 0.05 | fdr_ctrls < 0.05) %>%
    mutate(
      sig_group = case_when(
        fdr_cases < 0.05 & fdr_ctrls < 0.05 ~ "Both",
        fdr_cases < 0.05                     ~ "Cases only",
        fdr_ctrls < 0.05                     ~ "Controls only",
        TRUE                                  ~ "Neither"
      )
    )

  cat(sprintf("FDR<0.05 in either group: %s\n",
      format(nrow(sig_either), big.mark=",")))
  print(table(sig_either$sig_group))

  ggplot(sig_either,
         aes(x=beta_cases, y=beta_ctrls, color=sig_group)) +
    geom_point(alpha=0.5, size=1.2) +
    geom_abline(slope=1, intercept=0,
                linetype="dashed", color="grey40") +
    geom_hline(yintercept=0, color="grey70") +
    geom_vline(xintercept=0, color="grey70") +
    scale_color_manual(values=c(
      "Both"          = "#4DAF4A",
      "Cases only"    = "#E41A1C",
      "Controls only" = "#377EB8"
    )) +
    labs(title   = "Cases vs Controls Meta-mQTL Effect Sizes",
         subtitle= "FDR < 0.05 in at least one group",
         x       = "β (Cases meta-analysis)",
         y       = "β (Controls meta-analysis)",
         color   = NULL) +
    theme_minimal(base_size=13) +
    theme(legend.position="bottom")
}
```

## 5.4b Volcano Plots

```{r volcano-plots, fig.width=13, fig.height=5}
volcano_plots <- imap(meta_results, function(dat, group) {
  if (is.null(dat) || nrow(dat) == 0) return(NULL)

  # Significance thresholds
  fdr_thresh  <- dat %>% filter(FDR_BH < 0.05) %>%
    pull(meta_pval) %>% max(na.rm=TRUE)

  dat_plot <- dat %>%
    mutate(
      neg_log10_p = -log10(meta_pval),
      sig = case_when(
        FDR_BH < 0.05 & meta_beta > 0 ~ "FDR<0.05 (positive)",
        FDR_BH < 0.05 & meta_beta < 0 ~ "FDR<0.05 (negative)",
        TRUE                           ~ "Not significant"
      )
    )

  n_sig <- sum(dat_plot$sig != "Not significant", na.rm=TRUE)
  n_pos <- sum(dat_plot$sig == "FDR<0.05 (positive)", na.rm=TRUE)
  n_neg <- sum(dat_plot$sig == "FDR<0.05 (negative)", na.rm=TRUE)

  sig_colors <- c(
    "FDR<0.05 (positive)" = "#E41A1C",
    "FDR<0.05 (negative)" = "#377EB8",
    "Not significant"     = "grey70"
  )

  ggplot(dat_plot, aes(x=meta_beta, y=neg_log10_p, color=sig)) +
    geom_point(size=0.7, alpha=0.5) +
    geom_hline(yintercept=-log10(fdr_thresh),
               linetype="dashed", color="grey30", linewidth=0.5) +
    geom_vline(xintercept=0,
               linetype="dashed", color="grey30", linewidth=0.5) +
    scale_color_manual(values=sig_colors) +
    annotate("text", x=Inf, y=-log10(fdr_thresh),
             label="FDR=0.05", hjust=1.1, vjust=-0.5,
             size=3, color="grey30") +
    annotate("text", x=Inf, y=Inf,
             label=sprintf("FDR<0.05: %s\n(+) %s  (–) %s",
                           format(n_sig, big.mark=","),
                           format(n_pos, big.mark=","),
                           format(n_neg, big.mark=",")),
             hjust=1.05, vjust=1.3, size=3.2, color="grey20") +
    labs(
      title    = sprintf("%s Meta-Analysis Volcano Plot",
                         str_to_title(group)),
      subtitle = sprintf("%s total pairs",
                         format(nrow(dat), big.mark=",")),
      x        = "Meta-analysis β (effect size)",
      y        = "-log₁₀(p)",
      color    = NULL
    ) +
    theme_minimal(base_size=12) +
    theme(legend.position="bottom",
          legend.text=element_text(size=9))
})

volcano_plots <- Filter(Negate(is.null), volcano_plots)
if (length(volcano_plots) > 0)
  wrap_plots(volcano_plots, ncol=2) +
    plot_annotation(
      title = "Volcano Plots — Meta-Analysis Results",
      theme = theme(plot.title=element_text(size=15, face="bold"))
    )
```

## 5.5 Heterogeneity Summary

```{r heterogeneity}
for (group in names(meta_results)) {
  dat <- meta_results[[group]]
  if (is.null(dat)) next

  sig <- dat %>% filter(FDR_BH < 0.05)
  if (nrow(sig) == 0) next

  cat(sprintf("\n%s — heterogeneity among FDR<0.05 hits:\n",
      str_to_title(group)))
  cat(sprintf("  Median I²:              %.1f%%\n",
      median(sig$I2,    na.rm=TRUE)))
  cat(sprintf("  %% with Q p<0.05:       %.1f%%\n",
      100 * mean(sig$Q_pval < 0.05, na.rm=TRUE)))
  cat(sprintf("  %% with I² > 50%%:       %.1f%%\n",
      100 * mean(sig$I2 > 50, na.rm=TRUE)))

  # I² distribution plot
  p <- ggplot(sig, aes(x=I2)) +
    geom_histogram(bins=40, fill="steelblue", color="white") +
    geom_vline(xintercept=c(25,50,75),
               linetype="dashed", color="firebrick", alpha=0.6) +
    annotate("text", x=c(25,50,75), y=Inf,
             label=c("Low","Moderate","High"),
             vjust=2, hjust=-0.1, size=3, color="firebrick") +
    labs(title   = sprintf("%s: I² Distribution (FDR<0.05 hits)",
                            str_to_title(group)),
         x="I² (%)", y="Count") +
    theme_minimal(base_size=12)
  print(p)
}
```

---

# 6. Save Significant Results

```{r save-results}
for (group in names(meta_results)) {
  dat <- meta_results[[group]]
  if (is.null(dat)) next

  sig <- dat %>% filter(FDR_BH < 0.05)
  out <- file.path(META_DIR,
                    paste0(group, "_sig_fdr05.csv"))
  write.csv(sig, out, row.names=FALSE, quote=FALSE)
  cat(sprintf("%s: %s significant pairs (FDR<0.05) saved to %s\n",
      str_to_title(group),
      format(nrow(sig), big.mark=","),
      basename(out)))
}
```

---

# 7. Session Info

```{r session-info}
sessionInfo()
```
