---
title: "PPMI-PEG mQTL / metQTL Overlap Analysis"
subtitle: "Longitudinal mQTL meta-hits overlapped with PEG1 metabolomics QTLs, annotated via Ensembl"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = TRUE, message = FALSE)
library(dplyr)
library(readr)
library(tidyr)
library(stringr)
library(ggplot2)
library(httr)
library(jsonlite)
library(purrr)
library(knitr)

RESULTS_DIR <- "../../results/META_MQTL"

# Row cap for the metQTL trans files. Mirrors `downstream.max_trans_rows` in
# pipelines/PEG_METAB/config.yaml — kept as a literal here so the report stays
# readable standalone; update both together if it changes.
MAX_TRANS_ROWS <- 10000000
```

## Overview

This report combines four analysis stages:

1. **metQTL FDR handling and overlap** — combining PEG1 c18 + hilic trans-metQTL
   results, and overlapping FDR-significant SNP-metabolite pairs against the
   classified longitudinal PPMI mQTL results (from the meta-analysis hit list,
   re-tested with a mixed model across follow-up visits).
2. **Gene annotation** — resolving the overlapping SNPs to nearest/overlapping
   genes via the Ensembl REST API, since a shared SNP driving both a
   methylation and a metabolite association is the strongest signal of a
   pleiotropic locus worth following up.
3. **PD GWAS overlap and union table** — checking whether the case
   meta-analysis mQTL hits fall near a known PD risk locus from Nalls et al.
   (2019, Table S2), then building a union table of case hits supported by
   *either* the metQTL overlap *or* GWAS proximity (or both).
4. **Metabolite compound annotation** — resolving as many of the overlap
   metabolite features as possible from their raw `mz_rt` labels to putative
   compound identities, via xmsannotator mass/retention-time matching against
   HMDB, KEGG, and LipidMaps (see Step 6).

Metabolite features (`C18_mz_rt_...` / `HILIC_mz_rt_...`) are carried under
their original mass-spec feature labels through Steps 1-5. No collaborator
compound-ID mapping table ever materialized, so Step 6 instead resolves them
independently via xmsannotator mass/retention-time matching — with partial
coverage (23/33 features get any match, 11/33 a high-confidence named
compound; see Step 6 for caveats).

**A methodological note carried through this analysis:** the
`trans_metqtls.txt.gz` files are generated with a liberal `pvOutputThreshold`
(1e-2, set in `pipelines/PEG_METAB/config.yaml`) so that enough candidate SNPs
survive to make the overlap with the longitudinal mQTL hits informative. The
complete output is written gzipped and kept intact; the truncation to the first
`r format(MAX_TRANS_ROWS, big.mark = ",")` rows now happens **at read time**
here, by streaming the gzip and stopping at the cap, rather than by a manual
`head` on the cluster before transfer.

Two consequences follow, and both matter:

- The rows we load are *not* the full test space, so a BH correction recomputed
  on them alone would be invalid — it would rank an already-preselected low-p
  subset against itself instead of against the true, much larger, denominator
  of all tests performed. We therefore use **MatrixEQTL's own per-file FDR
  column**, which is computed against that true denominator before any
  truncation, rather than recomputing it here. This requires
  `noFDRsaveMemory = FALSE` in `run_metqtl.R`; with it on, MatrixEQTL omits the
  FDR column entirely.
- The cap is only safe because MatrixEQTL sorts its output by p-value ascending
  when it computes FDR. Truncating therefore discards the least significant
  tail, not an arbitrary slice.

c18 and hilic are treated as separate multiple-testing families rather than one
merged correction, since a combined correction isn't recoverable from truncated
data anyway.

---

## Step 1: Combine metQTL results and apply FDR filtering

```{r combine-metqtl, cache=TRUE, cache.lazy=FALSE}
combine_and_fdr <- function(group_label, files, fdr_threshold = 0.05) {

  combined <- bind_rows(lapply(files, function(f) {
    # read_tsv streams the gzip rather than expanding it, and n_max stops the
    # read at the cap, so peak memory is bounded by the cap and not by the
    # on-disk file — which at pval_trans = 1e-2 is several GB compressed.
    read_tsv(f, col_types = cols(SNP = col_character(), gene = col_character(),
                                  beta = col_double(), `t-stat` = col_double(),
                                  `p-value` = col_double(), FDR = col_double()),
              n_max = MAX_TRANS_ROWS, guess_max = Inf) %>%
      mutate(source_file = basename(dirname(f)))
  }))

  sig <- combined %>% filter(FDR < fdr_threshold)

  out_all <- file.path(RESULTS_DIR, sprintf("peg1_%s_metqtl_combined_fdr.csv", group_label))
  out_sig <- file.path(RESULTS_DIR, sprintf("peg1_%s_metqtl_sig_fdr05.csv", group_label))
  write_csv(combined, out_all)
  write_csv(sig, out_sig)

  list(combined = combined, sig = sig)
}

cases_metqtl    <- combine_and_fdr(
  "cases",
  c("../../results/PEG_METAB/metqtl/c18_peg1_cases/trans_metqtls.txt.gz",
    "../../results/PEG_METAB/metqtl/hilic_peg1_cases/trans_metqtls.txt.gz")
)

controls_metqtl <- combine_and_fdr(
  "controls",
  c("../../results/PEG_METAB/metqtl/c18_peg1_controls/trans_metqtls.txt.gz",
    "../../results/PEG_METAB/metqtl/hilic_peg1_controls/trans_metqtls.txt.gz")
)

metqtl_cases_sig    <- cases_metqtl$sig
metqtl_controls_sig <- controls_metqtl$sig
```

```{r metqtl-summary}
metqtl_summary <- tibble(
  group           = c("Cases", "Controls"),
  total_tests     = c(nrow(cases_metqtl$combined), nrow(controls_metqtl$combined)),
  fdr_sig_pairs   = c(nrow(metqtl_cases_sig), nrow(metqtl_controls_sig))
)
kable(metqtl_summary, caption = "PEG1 metQTL: total tests combined (c18+hilic) and FDR<0.05 pairs")
```

---

## Step 2: Overlap with classified longitudinal mQTL results

Overlap is defined by shared SNP identity, cases-vs-cases and
controls-vs-controls kept separate (different populations, not
directly comparable). A single pleiotropic SNP can legitimately show up with
multiple significant CpGs *and* multiple significant metabolites — that
many-to-many structure is expected biology, not a join error.

```{r overlap-mqtl-metqtl, cache=TRUE}
overlap_group <- function(group_label, mqtl_classified_file, metqtl_sig) {

  mqtl <- read_csv(mqtl_classified_file,
                    col_types = cols(pair_id = col_character(), snp_id = col_character(),
                                      cpg_id = col_character(), status = col_character(),
                                      warnings = col_character(), category = col_character(),
                                      same_sign_as_meta = col_logical(), replicated = col_logical(),
                                      dynamic = col_logical(), .default = col_double()),
                    guess_max = Inf)

  overlap <- mqtl %>%
    inner_join(metqtl_sig, by = c("snp_id" = "SNP"), relationship = "many-to-many") %>%
    rename(metabolite_feature = gene, metqtl_beta = beta, metqtl_pval = `p-value`, metqtl_fdr = FDR)

  write_csv(overlap, file.path(RESULTS_DIR, sprintf("mqtl_metqtl_overlap_%s.csv", group_label)))

  # Carry the mQTL denominator with the result so the summary table can report
  # it instead of a hardcoded literal that goes stale on the next re-run.
  attr(overlap, "n_mqtl_pairs") <- nrow(mqtl)
  overlap
}

overlap_cases <- overlap_group("cases",
  "../../results/PPMI_LONG/longitudinal_mqtl_cases_classified.csv", metqtl_cases_sig)

overlap_controls <- overlap_group("controls",
  "../../results/PPMI_LONG/longitudinal_mqtl_controls_classified.csv", metqtl_controls_sig)
```

```{r overlap-summary}
overlap_summary <- tibble(
  group                = c("Cases", "Controls"),
  # Derived, not hardcoded: these were literals (18986, 15282) from a run two
  # hit-lists ago and silently misreported the denominator once PPMI_LONG was
  # re-run against a new meta-analysis.
  mqtl_pairs_tested     = c(attr(overlap_cases, "n_mqtl_pairs"),
                            attr(overlap_controls, "n_mqtl_pairs")),
  metqtl_sig_pairs       = c(nrow(metqtl_cases_sig), nrow(metqtl_controls_sig)),
  unique_shared_snps     = c(n_distinct(overlap_cases$snp_id), n_distinct(overlap_controls$snp_id)),
  overlap_rows           = c(nrow(overlap_cases), nrow(overlap_controls))
)
kable(overlap_summary, caption = "mQTL x metQTL overlap summary")
```

### Longitudinal category breakdown among overlapping SNPs

```{r category-breakdown}
cat_cases <- overlap_cases %>% distinct(snp_id, category) %>% count(category) %>% mutate(group = "Cases")
cat_controls <- overlap_controls %>% distinct(snp_id, category) %>% count(category) %>% mutate(group = "Controls")
bind_rows(cat_cases, cat_controls) %>%
  pivot_wider(names_from = group, values_from = n, values_fill = 0) %>%
  kable(caption = "Longitudinal classification of SNPs shared with metQTL hits")
```

### Dynamic mQTLs (cases): metQTL overlap status

The overlap above is an *inner join* — it only shows hits that overlapped a
metQTL, so a longitudinally "dynamic" mQTL with no metabolite signal would be
invisible by construction. To see all dynamic hits regardless of metQTL
status, we take the classified longitudinal results (cases) as the **left**
table and left-join the metQTL significant set, keeping every dynamic hit
whether or not it matched.

```{r dynamic-metqtl-status}
cases_classified_full <- read_csv("../../results/PPMI_LONG/longitudinal_mqtl_cases_classified.csv",
                                   col_types = cols(pair_id = col_character(), snp_id = col_character(),
                                                     cpg_id = col_character(), status = col_character(),
                                                     warnings = col_character(), category = col_character(),
                                                     same_sign_as_meta = col_logical(), replicated = col_logical(),
                                                     dynamic = col_logical(), .default = col_double()),
                                   guess_max = Inf)

dynamic_hits <- cases_classified_full %>%
  filter(category == "replicated_and_dynamic") %>%
  left_join(metqtl_cases_sig %>% select(SNP, gene, `p-value`, FDR),
            by = c("snp_id" = "SNP"), relationship = "many-to-many") %>%
  rename(metabolite_feature = gene, metqtl_pval = `p-value`, metqtl_fdr = FDR) %>%
  mutate(has_metqtl_match = !is.na(metabolite_feature))

kable(dynamic_hits %>%
        select(snp_id, cpg_id, main_beta, main_fdr, interaction_p, interaction_fdr,
               has_metqtl_match, metabolite_feature, metqtl_fdr),
      caption = "All 'replicated_and_dynamic' case mQTLs, with metQTL match status (left join — unmatched rows show NA)")

no_metqtl <- dynamic_hits %>% filter(!has_metqtl_match)
cat("Dynamic mQTLs (cases) with NO metQTL overlap:", n_distinct(no_metqtl$snp_id), "\n")
```

```{r dynamic-no-overlap-table}
kable(no_metqtl %>% select(snp_id, cpg_id, main_beta, main_fdr, interaction_p, interaction_fdr),
      caption = "Dynamic mQTLs (cases) with no corresponding metabolite QTL signal")
```

---

## Step 3: Gene annotation via Ensembl REST API

```{r annotate-snps, cache=TRUE}
unique_snps <- union(overlap_cases$snp_id, overlap_controls$snp_id)

WINDOW <- 250000
BASE_URL <- "https://rest.ensembl.org"

annotate_snp <- function(snp_id) {
  parts <- strsplit(snp_id, ":")[[1]]
  chr <- parts[1]
  pos <- as.integer(parts[2])

  query_region <- function(start, end) {
    url <- sprintf("%s/overlap/region/human/%s:%d-%d?feature=gene;content-type=application/json",
                    BASE_URL, chr, start, end)
    resp <- tryCatch(GET(url), error = function(e) NULL)
    if (is.null(resp) || status_code(resp) != 200) return(NULL)
    fromJSON(content(resp, as = "text", encoding = "UTF-8"), flatten = TRUE)
  }

  hit <- query_region(pos, pos)
  Sys.sleep(0.1)

  if (!is.null(hit) && length(hit) > 0 && is.data.frame(hit) && nrow(hit) > 0) {
    gene_labels <- ifelse(is.na(hit$external_name) | hit$external_name == "",
                           hit$id, hit$external_name)
    genes <- unique(gene_labels)
    return(tibble(snp_id = snp_id, gene = paste(genes, collapse = ";"),
                   relationship = "overlapping", distance_bp = 0))
  }

  hit_wide <- query_region(max(1, pos - WINDOW), pos + WINDOW)
  Sys.sleep(0.1)

  if (is.null(hit_wide) || length(hit_wide) == 0 || !is.data.frame(hit_wide) || nrow(hit_wide) == 0) {
    return(tibble(snp_id = snp_id, gene = NA_character_,
                   relationship = paste0("none_within_", WINDOW, "bp"), distance_bp = NA_integer_))
  }

  hit_wide <- hit_wide %>%
    mutate(dist = pmax(0, pmax(start - pos, pos - end)),
           gene_label = ifelse(is.na(external_name) | external_name == "", id, external_name))

  nearest <- hit_wide %>% slice_min(dist, n = 1, with_ties = FALSE)

  tibble(snp_id = snp_id, gene = nearest$gene_label,
         relationship = "nearest", distance_bp = nearest$dist)
}

annotations <- map_dfr(unique_snps, function(s) {
  tryCatch(annotate_snp(s), error = function(e) {
    tibble(snp_id = s, gene = NA_character_, relationship = "query_failed", distance_bp = NA_integer_)
  })
})

write_csv(annotations, file.path(RESULTS_DIR, "overlap_snp_gene_annotations.csv"))

overlap_cases_annot <- overlap_cases %>%
  left_join(annotations, by = "snp_id") %>%
  rename(nearest_gene = gene, gene_relationship = relationship, gene_distance_bp = distance_bp)

overlap_controls_annot <- overlap_controls %>%
  left_join(annotations, by = "snp_id") %>%
  rename(nearest_gene = gene, gene_relationship = relationship, gene_distance_bp = distance_bp)

write_csv(overlap_cases_annot, file.path(RESULTS_DIR, "mqtl_metqtl_overlap_cases_annotated.csv"))
write_csv(overlap_controls_annot, file.path(RESULTS_DIR, "mqtl_metqtl_overlap_controls_annotated.csv"))
```

### Annotation outcome

```{r annotation-outcome}
kable(as.data.frame(table(relationship = annotations$relationship)),
      caption = "SNP -> gene annotation outcome (overlapping gene body, nearest within 250kb, or none found)")
```

### Genes implicated, cases

```{r genes-cases}
genes_cases <- overlap_cases_annot %>%
  distinct(snp_id, nearest_gene) %>%
  filter(!is.na(nearest_gene), nearest_gene != "") %>%
  count(nearest_gene, sort = TRUE)

kable(genes_cases, caption = "Unique genes implicated by shared SNPs — Cases")
```

### Genes implicated, controls

```{r genes-controls}
genes_controls <- overlap_controls_annot %>%
  distinct(snp_id, nearest_gene) %>%
  filter(!is.na(nearest_gene), nearest_gene != "") %>%
  count(nearest_gene, sort = TRUE)

kable(genes_controls, caption = "Unique genes implicated by shared SNPs — Controls")
```

### Gene frequency, top hits

```{r gene-plot, fig.width=9, fig.height=5}
top_genes <- bind_rows(
  genes_cases %>% mutate(group = "Cases"),
  genes_controls %>% mutate(group = "Controls")
) %>%
  group_by(group) %>%
  slice_max(n, n = 10) %>%
  ungroup()

ggplot(top_genes, aes(x = reorder(nearest_gene, n), y = n, fill = group)) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  facet_wrap(~group, scales = "free") +
  labs(x = NULL, y = "Number of shared SNPs",
       title = "Top genes implicated by mQTL x metQTL overlap SNPs") +
  theme_minimal(base_size = 12)
```

---

## Step 4: Overlap with Nalls et al. PD GWAS loci

Using the curated Nalls et al. (2019) locus table (Table S2) — already
restricted to genome-wide significant, LD-independent lead SNPs — rather than
the looser nominal-significance (p<0.05) SNP list, since the latter would
flag a large fraction of any SNP set as a "GWAS hit" by chance alone.

Because these are independent lead SNPs from a clumping procedure, they are
unlikely to be the exact variant present in the PPMI/PEG genotype panel. We
therefore use a **window-based match (±250kb)** rather than requiring
identical `chr:pos` — the same logic used for the Ensembl gene-annotation
step above. This window is a judgment call; tighten or loosen
`GWAS_WINDOW` below as appropriate.

```{r load-nalls, cache=TRUE}
GWAS_WINDOW <- 250000   # bp; matches the Ensembl gene-annotation window

nalls_raw <- read_tsv("../../Downloads/reference/Nalls_GWAS/PDNallsP005_TableS2.csv",
                       col_types = cols(.default = col_character()), guess_max = Inf)

qtl_gene_col   <- grep("QTL_Nominated_Gene", names(nalls_raw), value = TRUE)[1]
p_all_col      <- grep("^P_all_studies$", names(nalls_raw), value = TRUE)[1]
locus_num_col  <- grep("Locus_Number", names(nalls_raw), value = TRUE)[1]

nalls <- nalls_raw %>%
  transmute(
    nalls_snp     = SNP,
    CHR           = as.integer(CHR),
    BP            = as.integer(gsub(",", "", BP)),
    nearest_gene  = Nearest_Gene,
    qtl_gene      = .data[[qtl_gene_col]],
    p_all_studies = as.numeric(.data[[p_all_col]]),
    locus_number  = as.integer(.data[[locus_num_col]])
  )

stopifnot("Nalls BP parsing failed" = sum(is.na(nalls$BP)) == 0)
stopifnot("Nalls CHR parsing failed" = sum(is.na(nalls$CHR)) == 0)

cat("Nalls loci loaded:", nrow(nalls), "\n")
```

```{r match-gwas, cache=TRUE}
cases_meta <- read_csv("../../results/PPMI_LONG/longitudinal_mqtl_cases_classified.csv",
                        col_types = cols(pair_id = col_character(), snp_id = col_character(),
                                          cpg_id = col_character(), status = col_character(),
                                          warnings = col_character(), category = col_character(),
                                          same_sign_as_meta = col_logical(), replicated = col_logical(),
                                          dynamic = col_logical(), .default = col_double()),
                        guess_max = Inf)

case_snps <- cases_meta %>%
  distinct(snp_id) %>%
  mutate(chr = str_extract(snp_id, "^[^:]+"),
         pos = as.integer(str_extract(snp_id, "(?<=:)[0-9]+")))

match_gwas <- function(chr, pos) {
  cand <- nalls %>% filter(as.character(CHR) == chr)
  if (nrow(cand) == 0) {
    return(tibble(is_gwas_hit = FALSE, gwas_dist_bp = NA_integer_,
                   gwas_nearest_gene = NA_character_, gwas_locus_number = NA_integer_))
  }
  dists <- abs(cand$BP - pos)
  i <- which.min(dists)
  tibble(is_gwas_hit = dists[i] <= GWAS_WINDOW, gwas_dist_bp = dists[i],
         gwas_nearest_gene = cand$nearest_gene[i], gwas_locus_number = cand$locus_number[i])
}

gwas_matches <- case_snps %>%
  mutate(m = pmap(list(chr, pos), match_gwas)) %>%
  unnest(m)

cat("Unique case-hit SNPs tested against Nalls loci:", nrow(gwas_matches), "\n")
cat("Within", GWAS_WINDOW/1000, "kb of a known PD GWAS locus:", sum(gwas_matches$is_gwas_hit), "\n")
```

### GWAS-adjacent case hits

```{r gwas-hit-table}
gwas_hits_table <- gwas_matches %>%
  filter(is_gwas_hit) %>%
  arrange(gwas_dist_bp) %>%
  select(snp_id, gwas_nearest_gene, gwas_locus_number, gwas_dist_bp)

kable(gwas_hits_table, caption = "Case mQTL SNPs within 250kb of a known Nalls et al. PD GWAS locus")
```

---

## Step 5: Union table — cases supported by metQTL, GWAS, or both

```{r union-table, cache=TRUE}
metqtl_snp_set <- unique(metqtl_cases_sig$SNP)

union_table <- cases_meta %>%
  mutate(in_metqtl = snp_id %in% metqtl_snp_set) %>%
  left_join(gwas_matches %>% select(snp_id, is_gwas_hit, gwas_dist_bp,
                                     gwas_nearest_gene, gwas_locus_number),
            by = "snp_id") %>%
  rename(in_gwas = is_gwas_hit) %>%
  left_join(overlap_cases %>% distinct(snp_id, metabolite_feature),
            by = "snp_id", relationship = "many-to-many") %>%
  # Union criteria: metQTL support, GWAS proximity, OR longitudinal dynamic
  # classification — a dynamic hit is of interest in its own right even
  # without external metQTL/GWAS support, per the earlier left-join finding
  # that 2 dynamic case hits had no metQTL overlap.
  mutate(union_hit = in_metqtl | in_gwas | dynamic) %>%
  filter(union_hit)

write_csv(union_table, file.path(RESULTS_DIR, "cases_union_metqtl_or_gwas.csv"))

cat("Total case pairs in union (metQTL, GWAS, and/or dynamic):", nrow(union_table), "\n")
cat("Unique SNPs in union:", n_distinct(union_table$snp_id), "\n")
cat("  - metQTL only:            ", n_distinct(union_table$snp_id[union_table$in_metqtl & !union_table$in_gwas & !union_table$dynamic]), "\n")
cat("  - GWAS only:              ", n_distinct(union_table$snp_id[!union_table$in_metqtl & union_table$in_gwas & !union_table$dynamic]), "\n")
cat("  - Dynamic only (neither):  ", n_distinct(union_table$snp_id[!union_table$in_metqtl & !union_table$in_gwas & union_table$dynamic]), "\n")
cat("  - metQTL + GWAS:          ", n_distinct(union_table$snp_id[union_table$in_metqtl & union_table$in_gwas & !union_table$dynamic]), "\n")
cat("  - Any combo incl. dynamic:", n_distinct(union_table$snp_id[union_table$dynamic & (union_table$in_metqtl | union_table$in_gwas)]), "\n")
```

```{r union-summary-table}
union_table %>%
  distinct(snp_id, cpg_id, category, in_metqtl, in_gwas, dynamic,
           metabolite_feature, gwas_nearest_gene, gwas_dist_bp) %>%
  arrange(desc(dynamic), desc(in_metqtl & in_gwas)) %>%
  kable(caption = "Case meta-hits supported by metQTL overlap, GWAS proximity, and/or longitudinal dynamic classification")
```

---

## Step 6: Metabolite compound annotation (xmsannotator)

No collaborator compound-ID mapping table for the overlap metabolite
features ever materialized, so this step resolves them independently:
mass/retention-time matching (10ppm, ±30s) against the HMDB, KEGG, and
LipidMaps reference databases via
[xmsannotator](https://github.com/kuppal2/xmsannotator), run separately per
platform (`c18`/`hilic`) and database in
`annotate_overlap_metabolites_xmsannotator.R`. `c18`/`LipidMaps` did not
finish that run (only 2 initial DB matches, stopped short of a compound
call); the other 5 platform/db combinations completed. `rerun_step3_matching.R`
re-derives the consolidated match tables read in below from that run's
on-disk output without re-running the (expensive) annotation itself.

```{r metabolite-annotation, cache=TRUE}
best_annotation <- read_csv(file.path(RESULTS_DIR, "overlap_metabolite_best_annotation.csv"), show_col_types = FALSE) %>%
  select(metabolite_feature, annotated_compound, confidence, score, source)

n_any_match  <- n_distinct(best_annotation$metabolite_feature)
named_match  <- best_annotation %>% filter(confidence == 2, !annotated_compound %in% c("", "-"))

cat("Overlap metabolite features with any mz/rt match:      ", n_any_match, "/ 33\n")
cat("Overlap metabolite features with a high-confidence,\n",
    "named compound match:                                 ", nrow(named_match), "/ 33\n")

overlap_cases_compounds <- overlap_cases_annot %>%
  left_join(best_annotation, by = "metabolite_feature")

overlap_controls_compounds <- overlap_controls_annot %>%
  left_join(best_annotation, by = "metabolite_feature")

union_table <- union_table %>%
  left_join(best_annotation, by = "metabolite_feature")

write_csv(overlap_cases_compounds, file.path(RESULTS_DIR, "mqtl_metqtl_overlap_cases_annotated_compounds.csv"))
write_csv(overlap_controls_compounds, file.path(RESULTS_DIR, "mqtl_metqtl_overlap_controls_annotated_compounds.csv"))
write_csv(union_table, file.path(RESULTS_DIR, "cases_union_metqtl_or_gwas.csv"))
```

### Annotated compounds

A `confidence` of 2 / `score` of 100000 marks xmsannotator's high-confidence
named calls; `confidence` 0 entries are a raw mass/RT hit only (frequently
without a resolvable compound name) and are not treated as identifications
below.

```{r metabolite-annotation-table}
named_match %>%
  arrange(source, metabolite_feature) %>%
  select(metabolite_feature, annotated_compound, source) %>%
  kable(caption = "High-confidence (confidence = 2) compound matches for overlap metabolite features")
```

- The **GPX5/GPX6** metabolite feature (`C18_mz_rt_351.997_25.714`) got **no
  match** in either c18 table — it remains unresolved, so the
  glutathione/redox-pathway hypothesis raised earlier is still unconfirmed.
- The **ERLIN1** SNP's metabolite features fared better: two of its four
  associated features resolve to high-confidence matches, **N6-Methyl-L-lysine**
  and **Valyl-Valine** — both amino-acid derivatives, consistent with
  ERLIN1's role in ER lipid/protein handling but not a slam-dunk pathway
  link on their own.
- The **RUNX1** metabolite feature (`C18_mz_rt_709.527_218.804`) got no match.
- These are still single mass/RT hits within a fairly loose tolerance
  (10ppm/30s), not MS/MS-confirmed identities — treat names here as leads
  for literature/pathway follow-up, not confirmed compound IDs.
- The GWAS overlap (Step 4/5) is currently **cases only**, per the original
  request — extending the same window-match logic to the controls-side
  classified results is a straightforward addition if useful later.
- The `union_table` combines metQTL support, GWAS proximity, and longitudinal
  dynamic classification at the SNP level; a hit can qualify through any one
  criterion independently, and "any combo incl. dynamic" is the strongest
  converging evidence in this analysis — a locus independently implicated by
  PD genetic risk, metabolomics, methylation, and a time-varying genetic
  effect.
- The 2 `replicated_and_dynamic` case-level hits are now included in the
  union table regardless of metQTL/GWAS support (Step 5), and are shown
  explicitly in Step 2's dynamic-status table whether or not they overlap a
  metQTL.
- Full output files (`*_combined_fdr.csv`, `*_sig_fdr05.csv`,
  `mqtl_metqtl_overlap_*_annotated.csv`, `overlap_snp_gene_annotations.csv`,
  `cases_union_metqtl_or_gwas.csv`, `mqtl_metqtl_overlap_*_annotated_compounds.csv`)
  are written alongside this report for downstream use. The `*_annotated.csv`
  files (Step 3) stay gene-annotation-only, since they're also the input
  contract for `annotate_overlap_metabolites_xmsannotator.R` /
  `rerun_step3_matching.R`; the compound annotations from Step 6 are joined
  into separate `*_annotated_compounds.csv` files and into
  `cases_union_metqtl_or_gwas.csv` instead, to avoid changing that contract.
