---
title: "mQTL Enrichment Analysis: GO, KEGG & Reactome"
author: "Your Name"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  cache   = TRUE,
  fig.width  = 10,
  fig.height = 7
)
```

```{r libraries}
suppressPackageStartupMessages({
  library(tidyverse)
  library(data.table)
  library(missMethyl)       # methylation-aware GO/KEGG enrichment
  library(ReactomePA)       # Reactome pathway enrichment
  library(org.Hs.eg.db)    # human gene annotation
  library(IlluminaHumanMethylationEPICanno.ilm10b4.hg19)  # EPIC v1 annotation
  library(ggplot2)
  library(patchwork)
  library(scales)
})

# Explicitly prefer dplyr functions that are commonly masked
# by Bioconductor packages (matrixStats, AnnotationDbi etc.)
count    <- dplyr::count
select   <- dplyr::select
filter   <- dplyr::filter
rename   <- dplyr::rename
mutate   <- dplyr::mutate
```

---

# 1. Configuration

```{r config}
# ── Input files from meta-analysis pipeline ───────────────────────────────────
SIG_FILES <- list(
  cases    = "results/meta/cases_sig_fdr05.csv",
  controls = "results/meta/controls_sig_fdr05.csv"
)

# Full meta-analysis results (for background probe list)
META_FILES <- list(
  cases    = "results/meta/cases_meta_results.csv.gz",
  controls = "results/meta/controls_meta_results.csv.gz"
)

# Output directory
OUT_DIR <- "results/enrichment"
dir.create(OUT_DIR, recursive=TRUE, showWarnings=FALSE)

# FDR threshold for enrichment results
ENRICH_FDR <- 0.05

# Minimum gene set size (avoid very small/large gene sets)
MIN_GS_SIZE <- 10
MAX_GS_SIZE <- 500

# Colors
GROUP_COLORS <- c(cases="#E41A1C", controls="#377EB8")
```

---

# 2. Load Significant CpGs

```{r load-sig-cpgs}
# Load significant mQTL results per group
sig_list <- imap(SIG_FILES, function(path, group) {
  if (!file.exists(path)) {
    warning(sprintf("File not found: %s", path))
    return(NULL)
  }
  dat <- fread(path, data.table=FALSE)
  cat(sprintf("%-10s: %s significant pairs (FDR<0.05)\n",
      group, format(nrow(dat), big.mark=",")))
  dat
})
sig_list <- Filter(Negate(is.null), sig_list)

# Load full results for background CpG universe
meta_list <- imap(META_FILES, function(path, group) {
  if (!file.exists(path)) {
    warning(sprintf("Full results not found: %s", path))
    return(NULL)
  }
  dat <- fread(path, data.table=FALSE)
  cat(sprintf("%-10s: %s total tested pairs (background)\n",
      group, format(nrow(dat), big.mark=",")))
  dat
})
meta_list <- Filter(Negate(is.null), meta_list)
```

## 2.1 Extract Unique Significant CpGs

```{r extract-cpgs}
# For enrichment we work at the CpG level, not the SNP-CpG pair level.
# A CpG may appear multiple times (associated with different SNPs) —
# we deduplicate to get the unique set of CpGs under genetic regulation.

sig_cpgs <- imap(sig_list, function(dat, group) {
  cpgs <- unique(dat$cpg_id)
  cat(sprintf("%-10s: %s unique significant CpGs\n",
      group, format(length(cpgs), big.mark=",")))
  cpgs
})

# Background: all CpGs tested in the meta-analysis
bg_cpgs <- imap(meta_list, function(dat, group) {
  cpgs <- unique(dat$cpg_id)
  cat(sprintf("%-10s: %s background CpGs\n",
      group, format(length(cpgs), big.mark=",")))
  cpgs
})

# Summary
cat(sprintf("\nCases:    %s sig / %s background (%.1f%%)\n",
    format(length(sig_cpgs$cases),    big.mark=","),
    format(length(bg_cpgs$cases),     big.mark=","),
    100 * length(sig_cpgs$cases) / length(bg_cpgs$cases)))
cat(sprintf("Controls: %s sig / %s background (%.1f%%)\n",
    format(length(sig_cpgs$controls), big.mark=","),
    format(length(bg_cpgs$controls),  big.mark=","),
    100 * length(sig_cpgs$controls) / length(bg_cpgs$controls)))
```

---

# 3. EPIC v1 Annotation

```{r epic-annotation}
# Load EPIC v1 annotation — maps CpG IDs to genes and genomic features
anno <- as.data.frame(
  getAnnotation(IlluminaHumanMethylationEPICanno.ilm10b4.hg19)
)

cat(sprintf("EPIC v1 annotation: %d probes\n", nrow(anno)))
cat(sprintf("Annotation columns: %s\n",
    paste(colnames(anno)[1:10], collapse=", ")))

# Genomic feature distribution of significant CpGs
feature_dist <- imap_dfr(sig_cpgs, function(cpgs, group) {
  anno %>%
    filter(Name %in% cpgs) %>%
    dplyr::count(Relation_to_Island, UCSC_RefGene_Group) %>%
    mutate(group=group)
})

# CpG island relation distribution
imap_dfr(sig_cpgs, function(cpgs, group) {
  anno %>%
    filter(Name %in% cpgs) %>%
    dplyr::count(Relation_to_Island) %>%
    mutate(group=str_to_title(group),
           pct=round(100*n/sum(n), 1))
}) %>%
  pivot_wider(names_from=group, values_from=c(n, pct)) %>%
  knitr::kable(caption="CpG island relation of significant mQTL CpGs")
```

```{r genomic-feature-plot, fig.height=4}
imap_dfr(sig_cpgs, function(cpgs, group) {
  anno %>%
    filter(Name %in% cpgs) %>%
    mutate(
      feature = case_when(
        grepl("TSS",  UCSC_RefGene_Group) ~ "Promoter (TSS)",
        grepl("Body", UCSC_RefGene_Group) ~ "Gene Body",
        grepl("ExonBnd|3'UTR|5'UTR", UCSC_RefGene_Group) ~ "UTR/Exon",
        UCSC_RefGene_Group == "" ~ "Intergenic",
        TRUE ~ "Other"
      )
    ) %>%
    dplyr::count(feature) %>%
    mutate(group=str_to_title(group), pct=100*n/sum(n))
}) %>%
  ggplot(aes(x=reorder(feature, pct), y=pct, fill=group)) +
  geom_col(position="dodge", width=0.7) +
  scale_fill_manual(values=GROUP_COLORS) +
  coord_flip() +
  labs(title="Genomic Feature Distribution of Significant mQTL CpGs",
       x=NULL, y="Percentage (%)", fill=NULL) +
  theme_minimal(base_size=12) +
  theme(legend.position="bottom")
```

---

# 4. GO Enrichment (missMethyl — gometh)

`gometh()` tests GO term enrichment while correcting for the number of
CpG probes per gene — genes with more probes are more likely to appear
by chance in any CpG list, and this bias must be corrected.

```{r go-enrichment, cache=TRUE}
run_gometh <- function(sig_cpgs, bg_cpgs, collection="GO", group_name="") {
  cat(sprintf("\nRunning gometh (%s) for %s...\n", collection, group_name))

  result <- tryCatch(
    gometh(
      sig.cpg   = sig_cpgs,
      all.cpg   = bg_cpgs,
      collection = collection,
      array.type = "EPIC",
      prior.prob = TRUE   # use probe-count correction
    ),
    error = function(e) {
      cat(sprintf("  ERROR: %s\n", e$message))
      NULL
    }
  )

  if (is.null(result)) return(NULL)

  result <- result %>%
    filter(N >= MIN_GS_SIZE, N <= MAX_GS_SIZE) %>%
    arrange(FDR) %>%
    mutate(group=group_name)

  n_sig <- sum(result$FDR < ENRICH_FDR, na.rm=TRUE)
  cat(sprintf("  Significant terms (FDR<%s): %d\n", ENRICH_FDR, n_sig))
  result
}

# Run GO enrichment for both groups
go_results <- list()
for (group in names(sig_cpgs)) {
  go_results[[group]] <- run_gometh(
    sig_cpgs = sig_cpgs[[group]],
    bg_cpgs  = bg_cpgs[[group]],
    collection = "GO",
    group_name = group
  )
}
```

## 4.1 Top GO Terms

```{r go-top-terms}
for (group in names(go_results)) {
  res <- go_results[[group]]
  if (is.null(res)) next

  sig <- res %>% filter(FDR < ENRICH_FDR)
  cat(sprintf("\n### %s — %d significant GO terms\n",
      str_to_title(group), nrow(sig)))

  if (nrow(sig) == 0) {
    cat("No significant GO terms at FDR <", ENRICH_FDR, "\n")
    next
  }

  sig %>%
    slice_head(n=20) %>%
    select(ONTOLOGY, TERM, N, DE, P.DE, FDR) %>%
    mutate(
      P.DE = formatC(P.DE, format="e", digits=2),
      FDR  = formatC(FDR,  format="e", digits=2)
    ) %>%
    knitr::kable(
      col.names = c("Ontology", "Term", "N genes",
                    "N DE", "p-value", "FDR"),
      caption = sprintf("Top GO terms — %s", str_to_title(group))
    ) %>%
    print()
}
```

## 4.2 GO Bubble Plots

```{r go-bubble, fig.height=8}
plot_go_bubble <- function(res, group, n_top=20, title_suffix="",
                           label="Terms") {
  if (is.null(res)) return(NULL)

  # Normalise column names — GO uses TERM/ONTOLOGY/N/DE/FDR
  # KEGG uses TERM/N/DE/FDR (no ONTOLOGY column)
  # Rename to common names for plotting
  if ("TERM" %in% colnames(res) && !"term_label" %in% colnames(res)) {
    res <- res %>% dplyr::rename(term_label=TERM)
  } else if ("Description" %in% colnames(res)) {
    res <- res %>% dplyr::rename(term_label=Description)
  } else {
    warning("Cannot find term label column in enrichment results")
    return(NULL)
  }

  # Size column: DE for gometh, Count for ReactomePA
  size_col <- if ("DE" %in% colnames(res)) "DE" else "Count"

  # Color column: ONTOLOGY for GO, absent for KEGG/Reactome
  has_ontology <- "ONTOLOGY" %in% colnames(res)

  plot_dat <- res %>%
    filter(FDR < ENRICH_FDR) %>%
    slice_head(n=n_top) %>%
    dplyr::mutate(
      neg_log10_fdr = -log10(FDR),
      term_label    = str_trunc(term_label, 50),
      term_label    = reorder(term_label, neg_log10_fdr)
    )

  if (nrow(plot_dat) == 0) {
    return(ggplot() +
      annotate("text", x=0.5, y=0.5,
               label=sprintf("No significant %s", label)) +
      theme_void())
  }

  p <- ggplot(plot_dat,
         aes(x=neg_log10_fdr, y=term_label,
             size=.data[[size_col]])) +
    geom_vline(xintercept=-log10(ENRICH_FDR),
               linetype="dashed", color="grey50") +
    scale_size_continuous(name="Genes hit", range=c(3,10)) +
    labs(
      title = sprintf("%s%s — Top %s",
                      str_to_title(group), title_suffix, label),
      x = "-log₁₀(FDR)",
      y = NULL
    ) +
    theme_minimal(base_size=11) +
    theme(legend.position="right",
          axis.text.y=element_text(size=9))

  if (has_ontology) {
    p <- p +
      geom_point(aes(color=ONTOLOGY), alpha=0.8) +
      scale_color_brewer(palette="Set1", name="Ontology")
  } else {
    p <- p +
      geom_point(color="steelblue", alpha=0.8)
  }
  p
}

go_plots <- imap(go_results, plot_go_bubble)
go_plots <- Filter(Negate(is.null), go_plots)

if (length(go_plots) > 0)
  wrap_plots(go_plots, ncol=1) +
    plot_annotation(
      title = "GO Enrichment — Significant mQTL CpGs",
      theme = theme(plot.title=element_text(size=14, face="bold"))
    )
```

## 4.3 Cases vs Controls GO Comparison

```{r go-comparison}
# Find GO terms significant in both, cases only, or controls only
if (!is.null(go_results$cases) && !is.null(go_results$controls)) {

  cases_sig <- go_results$cases    %>%
    filter(FDR < ENRICH_FDR) %>% pull(TERM)
  ctrls_sig <- go_results$controls %>%
    filter(FDR < ENRICH_FDR) %>% pull(TERM)

  cat(sprintf("GO terms significant in cases only:    %d\n",
      length(setdiff(cases_sig, ctrls_sig))))
  cat(sprintf("GO terms significant in controls only: %d\n",
      length(setdiff(ctrls_sig, cases_sig))))
  cat(sprintf("GO terms significant in both:          %d\n",
      length(intersect(cases_sig, ctrls_sig))))

  if (length(intersect(cases_sig, ctrls_sig)) > 0) {
    cat("\nShared significant GO terms:\n")
    print(intersect(cases_sig, ctrls_sig))
  }
}
```

---

# 5. KEGG Enrichment (missMethyl)

```{r kegg-enrichment, cache=TRUE}
kegg_results <- list()
for (group in names(sig_cpgs)) {
  kegg_results[[group]] <- run_gometh(
    sig_cpgs   = sig_cpgs[[group]],
    bg_cpgs    = bg_cpgs[[group]],
    collection = "KEGG",
    group_name = group
  )
}
```

```{r kegg-top}
for (group in names(kegg_results)) {
  res <- kegg_results[[group]]
  if (is.null(res)) next

  sig <- res %>% filter(FDR < ENRICH_FDR)
  cat(sprintf("\n### %s — %d significant KEGG pathways\n",
      str_to_title(group), nrow(sig)))

  if (nrow(sig) == 0) {
    cat("No significant KEGG pathways at FDR <", ENRICH_FDR, "\n")
    next
  }

  sig %>%
    slice_head(n=20) %>%
    select(TERM, N, DE, P.DE, FDR) %>%
    mutate(
      P.DE = formatC(P.DE, format="e", digits=2),
      FDR  = formatC(FDR,  format="e", digits=2)
    ) %>%
    knitr::kable(
      col.names = c("Pathway", "N genes", "N DE", "p-value", "FDR"),
      caption   = sprintf("Top KEGG pathways — %s", str_to_title(group))
    ) %>%
    print()
}
```

```{r kegg-plot, fig.height=7}
kegg_plots <- imap(kegg_results, ~plot_go_bubble(.x, .y, n_top=20, label="KEGG Pathways"))
kegg_plots <- Filter(Negate(is.null), kegg_plots)

if (length(kegg_plots) > 0)
  wrap_plots(kegg_plots, ncol=1) +
    plot_annotation(
      title = "KEGG Enrichment — Significant mQTL CpGs",
      theme = theme(plot.title=element_text(size=14, face="bold"))
    )
```

---

# 6. Reactome Enrichment

Reactome enrichment uses Entrez gene IDs. We map CpGs to genes via
the EPIC annotation then run `enrichPathway()` from ReactomePA.

```{r cpg-to-entrez}
# Map CpGs to Entrez IDs via EPIC annotation
cpg_to_entrez <- function(cpg_ids) {
  anno %>%
    filter(Name %in% cpg_ids) %>%
    select(Name, UCSC_RefGene_Name) %>%
    separate_rows(UCSC_RefGene_Name, sep=";") %>%
    filter(UCSC_RefGene_Name != "") %>%
    distinct(UCSC_RefGene_Name) %>%
    pull(UCSC_RefGene_Name) %>%
    { suppressMessages(
        AnnotationDbi::select(
          org.Hs.eg.db,
          keys    = .,
          columns = "ENTREZID",
          keytype = "SYMBOL"
        )
      ) } %>%
    filter(!is.na(ENTREZID)) %>%
    pull(ENTREZID) %>%
    unique()
}

# Map significant and background CpGs to Entrez IDs
cat("Mapping CpGs to Entrez gene IDs...\n")

sig_entrez <- imap(sig_cpgs, function(cpgs, group) {
  ids <- cpg_to_entrez(cpgs)
  cat(sprintf("  %-10s sig:  %d CpGs -> %d genes\n",
      group, length(cpgs), length(ids)))
  ids
})

bg_entrez <- imap(bg_cpgs, function(cpgs, group) {
  ids <- cpg_to_entrez(cpgs)
  cat(sprintf("  %-10s bg:   %d CpGs -> %d genes\n",
      group, length(cpgs), length(ids)))
  ids
})
```

```{r reactome-enrichment, cache=TRUE}
run_reactome <- function(sig_ids, bg_ids, group_name) {
  cat(sprintf("\nRunning Reactome enrichment for %s...\n", group_name))

  if (length(sig_ids) == 0) {
    cat("  No genes to test\n")
    return(NULL)
  }

  result <- tryCatch(
    enrichPathway(
      gene          = sig_ids,
      universe      = bg_ids,
      organism      = "human",
      pAdjustMethod = "BH",
      pvalueCutoff  = 0.05,
      qvalueCutoff  = ENRICH_FDR,
      minGSSize     = MIN_GS_SIZE,
      maxGSSize     = MAX_GS_SIZE,
      readable      = TRUE   # convert Entrez IDs to gene symbols
    ),
    error = function(e) {
      cat(sprintf("  ERROR: %s\n", e$message))
      NULL
    }
  )

  if (is.null(result) || nrow(as.data.frame(result)) == 0) {
    cat("  No significant Reactome pathways\n")
    return(NULL)
  }

  res_df <- as.data.frame(result) %>%
    filter(p.adjust < ENRICH_FDR) %>%
    mutate(group=group_name)

  cat(sprintf("  Significant pathways (FDR<%s): %d\n",
      ENRICH_FDR, nrow(res_df)))
  list(result=result, df=res_df)
}

reactome_results <- list()
for (group in names(sig_entrez)) {
  reactome_results[[group]] <- run_reactome(
    sig_ids    = sig_entrez[[group]],
    bg_ids     = bg_entrez[[group]],
    group_name = group
  )
}
```

## 6.1 Top Reactome Pathways

```{r reactome-top}
for (group in names(reactome_results)) {
  res <- reactome_results[[group]]
  if (is.null(res)) next

  cat(sprintf("\n### %s — %d significant Reactome pathways\n",
      str_to_title(group), nrow(res$df)))

  res$df %>%
    slice_head(n=20) %>%
    select(Description, GeneRatio, BgRatio, p.adjust, geneID) %>%
    mutate(
      p.adjust = formatC(p.adjust, format="e", digits=2),
      # Truncate long gene lists
      geneID   = str_trunc(geneID, 60)
    ) %>%
    knitr::kable(
      col.names = c("Pathway", "Gene Ratio", "BG Ratio",
                    "FDR", "Genes"),
      caption   = sprintf("Top Reactome pathways — %s",
                          str_to_title(group))
    ) %>%
    print()
}
```

## 6.2 Reactome Dot Plots

```{r reactome-dotplot, fig.height=8}
reactome_plots <- imap(reactome_results, function(res, group) {
  if (is.null(res)) return(NULL)

  plot_dat <- res$df %>%
    slice_head(n=20) %>%
    mutate(
      Description   = str_trunc(Description, 50),
      Description   = reorder(Description, -p.adjust),
      neg_log10_fdr = -log10(p.adjust),
      # Parse GeneRatio to numeric
      gene_ratio_n  = as.numeric(sub("/.*", "", GeneRatio)),
      gene_ratio_d  = as.numeric(sub(".*/", "", GeneRatio)),
      gene_ratio    = gene_ratio_n / gene_ratio_d
    )

  if (nrow(plot_dat) == 0) return(NULL)

  ggplot(plot_dat,
         aes(x=gene_ratio, y=Description,
             size=gene_ratio_n, color=neg_log10_fdr)) +
    geom_point() +
    scale_color_gradient(low="steelblue", high="firebrick",
                         name="-log₁₀(FDR)") +
    scale_size_continuous(name="Genes hit", range=c(3,10)) +
    labs(
      title = sprintf("%s — Top Reactome Pathways",
                      str_to_title(group)),
      x = "Gene Ratio", y = NULL
    ) +
    theme_minimal(base_size=11) +
    theme(axis.text.y=element_text(size=9))
})
reactome_plots <- Filter(Negate(is.null), reactome_plots)

if (length(reactome_plots) > 0)
  wrap_plots(reactome_plots, ncol=1) +
    plot_annotation(
      title = "Reactome Enrichment — Significant mQTL CpGs",
      theme = theme(plot.title=element_text(size=14, face="bold"))
    )
```

## 6.3 Cases vs Controls Reactome Comparison

```{r reactome-comparison}
if (!is.null(reactome_results$cases) &&
    !is.null(reactome_results$controls)) {

  cases_paths <- reactome_results$cases$df$Description
  ctrls_paths <- reactome_results$controls$df$Description

  cat(sprintf("Reactome pathways — cases only:    %d\n",
      length(setdiff(cases_paths, ctrls_paths))))
  cat(sprintf("Reactome pathways — controls only: %d\n",
      length(setdiff(ctrls_paths, cases_paths))))
  cat(sprintf("Reactome pathways — both:          %d\n",
      length(intersect(cases_paths, ctrls_paths))))

  # Shared pathways table
  shared <- intersect(cases_paths, ctrls_paths)
  if (length(shared) > 0) {
    bind_rows(
      reactome_results$cases$df %>%
        filter(Description %in% shared) %>%
        select(Description, p.adjust) %>%
        rename(FDR_cases=p.adjust),
      reactome_results$controls$df %>%
        filter(Description %in% shared) %>%
        select(Description, p.adjust) %>%
        rename(FDR_controls=p.adjust)
    ) %>%
      group_by(Description) %>%
      summarise(
        FDR_cases    = first(FDR_cases,    na_rm=TRUE),
        FDR_controls = first(FDR_controls, na_rm=TRUE)
      ) %>%
      mutate(across(starts_with("FDR"),
                    ~formatC(.x, format="e", digits=2))) %>%
      knitr::kable(
        col.names = c("Pathway", "FDR (Cases)", "FDR (Controls)"),
        caption   = "Reactome pathways significant in both groups"
      )
  } else {
    cat("No pathways significant in both groups.\n")
  }
}
```

---

# 7. Save Results

```{r save-results}
for (group in names(go_results)) {
  if (!is.null(go_results[[group]]))
    write.csv(go_results[[group]] %>% filter(FDR < ENRICH_FDR),
              file.path(OUT_DIR, paste0(group, "_GO_sig.csv")),
              row.names=FALSE, quote=FALSE)
}

for (group in names(kegg_results)) {
  if (!is.null(kegg_results[[group]]))
    write.csv(kegg_results[[group]] %>% filter(FDR < ENRICH_FDR),
              file.path(OUT_DIR, paste0(group, "_KEGG_sig.csv")),
              row.names=FALSE, quote=FALSE)
}

for (group in names(reactome_results)) {
  if (!is.null(reactome_results[[group]]))
    write.csv(reactome_results[[group]]$df,
              file.path(OUT_DIR, paste0(group, "_Reactome_sig.csv")),
              row.names=FALSE, quote=FALSE)
}

cat("Saved enrichment results to:", OUT_DIR, "\n")
```

---

# 8. Session Info

```{r session-info}
sessionInfo()
```
